---
title: "Lecture 6: Bayesian Hypothesis Testing Continued"
author: "Merlise Clyde"
date: "September 14, 2021"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Hypothesis Testing Setup Recap


- univariate data $y_i \overset{iid}{\sim} \mathcal{N}(\theta, 1)$ 

--

- test $\mathcal{H}_0: \theta = 0; \ \ \text{vs } \mathcal{H}_1: \theta \neq 0$ 

--
1. Put a prior on $\theta \mid \cal{H}_i$, 
$$\pi(\theta \mid \cal{H}_1) = \mathcal{N}(0, 1/\tau_0)$$ $$p(\theta \mid \cal{H}_0) = \delta_o(\theta)$$

--
2. Put a prior on the hypotheses $\pi(\mathcal{H}_0)$ and $\pi(\mathcal{H}_1)$.
      
--
3. Likelihood of the hypotheses  $p(y^{(n)} \mid \cal{H}_i)$
   

--
4. Obtain posterior probabilities of $\cal{H}_0$ and $\cal{H}_1$ via Bayes Theorem or Bayes Factors

--
5. Report based on loss (optional)


---
## Bayes factors

- **Bayes factor**: is a ratio of marginal likelihoods and it provides a weight of evidence in the data in favor of one model over another.

--

- **Rule of thumb**: $\mathcal{BF}_{10} > 10$ is strong evidence for $\mathcal{H}_1$;  $\mathcal{BF}_{10} > 100$ is decisive evidence for $\mathcal{H}_1$.  

--
**Not worth a bare mention**
$1 < \mathcal{BF}_{10} < 3$  

--

- Posterior probabilities 
$$
\begin{split}
\pi(\mathcal{H}_1 \mid Y) = \frac{ 1 }{\frac{\pi(\cal{H}_0)}{\pi(\cal{H}_1)} \frac{p(y^{(n)} \mid \mathcal{H}_0)}{p(y^{(n)} \mid \mathcal{H}_1)} + 1 } = \frac{ 1 }{ \cal{O}_{01} \mathcal{BF}_{01} + 1 } \\
\end{split}
$$

-  $\cal{O}_{01}$ prior odds of $\cal{H}_0$ to $\cal{H}_1$
  
--

Alternative expression for Bayes Factor

$$\cal{BF}_{10} = \frac{p(y^{(n)} \mid \cal{H}_1)}
       {p(y^{(n)} \mid \cal{H}_0)} =
  \frac{\pi_\theta(0 \mid \cal{H}_1)} 
       {\pi_\theta(0 \mid  y^{(n)}, \cal{H}_1)}$$

  
---
##  Marginal Likelihoods & Evidence

```{r marglik, fig.height=4.5, echo=F}
n = 10; y = 1.96/sqrt(n)
th <- seq(y - 3/sqrt(n), y + 3/sqrt(n), length=1000)
v0 = 1; tau0 = 1/v0
marg.H1 = dnorm(0, y, sqrt(v0 + 1/n))
theta.n = n*y/(n + tau0)
tau.n = n + tau0
plot(th,dnorm(th,y,sqrt(1/n)),
      type="l", lwd=2, ylim=c(0, dnorm(theta.n, theta.n, sqrt(1/tau.n))),
      col="blue",xlab=expression(theta),
      ylab=expression(pi(theta)))
lines(th, dnorm(th, 0, sqrt(v0)), col="red", lwd=2)
marg.H1 = dnorm(0, y, sqrt(v0 + 1/n))
segments(0, 0, 0, marg.H1, lty=2, lwd=2, col="purple")
abline(h =marg.H1, lty=2, lwd=3, col="purple")
#m = integrate(f = function(x) {dbinom(y,n,x)}, low=0.00001, upper=.99999)
#abline(h=m$value, lty=4, lwd=2, col="red")
lines(th, dnorm(th, theta.n, sqrt(1/tau.n)), lty=3,  lwd=3, col="purple")
marg.H0 = dnorm(0, y, sqrt(1/n))
segments(0, 0, 0, marg.H0, lty=2, lwd=2, col="blue")
abline(h=dnorm(y,y, sqrt(1/n)), lty=3, lwd=2, col="blue")
bf10 = marg.H1/marg.H0
bf01 = 1/bf10
post.prob.H1 = 1/(1 + bf01)

#cand = dnorm(0,0, sqrt(v0))/dnorm(0, theta.n, sqrt(1/tau.n))
```

  $\cal{BF}_{10}$ = `r round(bf10, 2)`
  Posterior Probability of $\cal{H}_0$ = `r round(1 - post.prob.H1, 4)` versus p-value of $0.05$
---
## Decisions

- Selection 0-1 loss; 
   + if $\pi(\cal{H}_1 \mid y^{(n)}) > .5$ choose $\cal{H}_1$, 
   + otherwise $\cal{H}_0$

--

- Estimation of $\theta$ under squared error loss 

- report $\hat{\theta}$ that minimizes Bayes expected loss

$$
\textsf{E}_{\theta \mid y^{(n)}} \left[(\theta - \hat{\theta})^2\right]
$$

--

- Bayes optimal estimator under squared error is the posterior mean  $\textsf{E}[\theta \mid y^{(n)}]$

--

- no $\cal{H}_i$!

--

- marginal posterior distribution of $\theta$

---
## Averaging over Hypotheses

Prior on $\theta$ is a mixture model:

$$p(\theta) = \pi_0 \delta_0(\theta) + (1 - \pi_0) \pi(\theta \mid \cal{H}_1)$$
--

- Dirac delta - degenerate distribution at 0


--
- "spike & slab" prior

```{r, fig=TRUE, echo=FALSE, fig.height=4, fig.width=5,out.width="50%"}
th = seq(-3, 3, length=1000)
d = dnorm(th, 0, 1)
pi.0 = 0.65

d = (1 - pi.0)*d/dnorm(0,0,1)
plot(th, d, type="l", col = 1, 
     xlab=expression(theta),
     ylab = expression(p(theta)),
     ylim=c(0, pi.0))
segments(0, 0, 0, pi.0, lty=2)     
```


--

- how to sample from prior?

---
## Posterior under Spike & Slab Prior

$$\pi(\theta \mid y^{(n)}) =  \Pr( \cal{H}_0 \mid y^{(n)}) \pi(\theta \mid \cal{H}_0, y^{(n)}) + \Pr( \cal{H}_1 \mid y^{(n)}) \pi(\theta \mid \cal{H}_1, y^{(n)})$$
--

$$\pi(\theta \mid y^{(n)}) =  \Pr( \cal{H}_0 \mid y^{(n)}) \delta_0(\theta) + \Pr( \cal{H}_1 \mid y^{(n)}) \pi(\theta \mid \cal{H}_1, y^{(n)})$$

- posterior also has a spike & slab

--

- mixture weights are updated

--

- updated "slab" hyperparameters

```{r postBMA, fig.height=4.5, out.width="60%",echo=F}
n = 10; y = 1.96/sqrt(n)
th <- seq(y - 3/sqrt(n), y + 3/sqrt(n), length=1000)
v0 = 1; tau0 = 1/v0
marg.H1 = dnorm(0, y, sqrt(v0 + 1/n))
marg.H0 = dnorm(0, y, sqrt(1/n))
theta.n = n*y/(n + tau0)
tau.n = n + tau0
bf10 = marg.H1/marg.H0
bf01 = 1/bf10
post.prob.H1 = 1/(1 + bf01)
d.post = post.prob.H1*dnorm(th, theta.n, sqrt(1/tau.n))/
         dnorm(theta.n, theta.n, sqrt(1/tau.n))

plot(th, d.post,
      type="l", lwd=2,
      col=1,xlab=expression(theta),
      ylab=expression(pi(theta)))

segments(0, 0, 0, 1 - post.prob.H1, lty=2, lwd=2, col=1)
```

---
## Prior
An important issue with hypothesis testing and using Spike & Slab prior is choice of hyperparameters  ( $\tau_0$ in this case)

--
- Bayes Factor and posterior probabilities of $\cal{H}_i$ depend on $\tau_0$  through $p(y^{(n)} \mid \cal{H_1})$

--

1. What is impact of  $\tau_0$ on $\cal{BF}_{01}$ ?

--
2.  How do we choose $\tau_0$?

---
## Question 1.

$$\cal{BF}_{01} = \frac{\pi(0 \mid \cal{H}_1, y^{(n)})}{\pi(0 \mid \cal{H_1})}$$

---
## Precision

```{r marglik2, fig.height=4.5, echo=F, out.width="50%"}
v0 = 10; tau0 = 1/v0
marg.H1 = dnorm(0, y, sqrt(v0 + 1/n))
theta.n = n*y/(n + tau0)
tau.n = n + tau0
plot(th,dnorm(th,y,sqrt(1/n)),
      type="l", lwd=2, ylim=c(0, dnorm(theta.n, theta.n, sqrt(1/tau.n))),
      col="blue",xlab=expression(theta),
      ylab=expression(pi(theta)))
lines(th, dnorm(th, 0, sqrt(v0)), col="red", lwd=2)
marg.H1 = dnorm(0, y, sqrt(v0 + 1/n))
segments(0, 0, 0, marg.H1, lty=2, lwd=2, col="purple")
abline(h=marg.H1, lty=2, lwd=3, col="purple")
lines(th, dnorm(th, theta.n, sqrt(1/tau.n)), lty=3,  lwd=3, col="purple")
marg.H0 = dnorm(0, y, sqrt(1/n))
segments(0, 0, 0, marg.H0, lty=2, lwd=2, col="blue")
abline(h=dnorm(y,y, sqrt(1/n)), lty=3, lwd=2, col="blue")
bf10 = marg.H1/marg.H0
bf01 = 1/bf10
post.prob.H1 = 1/(1 + bf01)
```

- $\tau_0 = 1/`r v0`$

--
- Bayes Factor for $\cal{H}_0$ to $\cal{H}_1$ is $`r round(bf01, 2)`$

--

- Posterior Probability of $\cal{H}_0$ = `r round(1 - post.prob.H1, 4)`

What about even more vague priors?
---
## Vague Priors & Hypothesis Testing



- As $\tau_0 \to 0$ the $\cal{BF}_{01} \to \infty$ and  $\Pr(\cal{H}_0 \mid y^{(n)}) \to 1$! 

--

- As we use a less & less informative prior under $\cal{H}_1$ we obtain more & more evidence for $\cal{H}_0$ over $\cal{H}_1$!

--

Known as **Bartlett's Paradox** - the paradox is that a seemingly non-informative prior for $\theta$ is very informative about $\cal{H}$!

--

- General problem with nested sequence of models.  If we choose vague priors on the additional parameter in the larger model we will be favoring the smaller models under consideration!

--

**Bottom Line** Don't use vague priors!

--

--

What then?

---
## Objective Bayes

- Conventional Priors

--

- Simplest is Unit Information Prior  (UIP)

--

- center prior at MLE ( $\bar{y}$ ) but choose prior precision to be the equivalent of a sample size of 1

--

- center prior at 0,  but choose prior precision to be the equivalent of a sample size of 1  (UIP)

--

Default UIP
$$\theta  \mid\cal{H}_1 \sim \textsf{N}(0, 1)$$

---
## UIP & BIC

Note:  UIP is the basis for the Bayes Information Criterion (BIC)

--

- BIC is derived in more general settings by taking a Laplace approximation to the marginal likelihood and making some simplifying assumptions
   
--
  
- BIC  chooses model with highest marginal likelihood
   
--

- BIC has a well known tendency to choose/favor simpler models due to UIP containing very little information
   
--

- consistent for model selection i.e. $\Pr(\cal{H}_i \mid y^{(n)})$ goes to 1 for the true model as $n \to \infty$


--

-  Is a fixed $\tau_0$ consistent as $n \to \infty$?

---
## Other Options

- Place a prior on $\tau_0$

--

$$\tau_0 \sim \textsf{Gamma}(1/2, 1/2)$$
$$p(\tau_0 \mid \cal{H}_1) = \frac{(1/2)^{1/2}}{\Gamma(1/2)} \tau_0^{1/2 -1} \exp{(-\tau_0/2)}$$
--

- If $\theta \mid \tau_0, \cal{H}_1 \sim \textsf{N}(0, 1/\tau_0)$,  then $\theta_0  \mid \cal{H}_1$ has a $\textsf{Cauchy}(0,1)$ distribution!  Recommended by Jeffreys (1961)

--

- no closed form expressions for marginal likelihood!




---
## Intrinsic Bayes Factors & Priors  (Berger & Pericchi)

- Can't use improper priors under $\cal{H}_1$

--

-  use part of the data $y(l)$ to update an improper prior on $\theta$ to get a proper posterior  $\pi(\theta \mid \cal{H}_i, y(l))$

--

- use $\pi(\theta \mid y(l), \cal{H}_i)$ to obtain the posterior for $\theta$ based on the rest of the training data

--

- Calculate a Bayes Factor (avoids arbitrary normalizing constants!)

--

- Choice of training sample $y(l)$? 

- Berger & Pericchi (1996) propose "averaging" over training samples  **intrinsic Bayes Factors**


--

- **intrinsic prior** on $\theta$ that leads to the IBF  


