---
title: "Lecture 8: Metropolis Algorithms and Diagnostics"
author: "Merlise Clyde"
date: "September 21"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(lattice)
require(MCMCpack)
require(hdrcde)
require(coda)
#require(rethinking)
set.seed(42)
```



## Last Class: Normal Means Model

Data Model
$$Y_i \mid \mu_i, \sigma^2 \overset{iid}{\sim} \textsf(\mu_i, \sigma^2)$$


Means Model 
$$\mu_i \mid \mu, \sigma^2_\mu \overset{iid}{\sim} \textsf(\mu, \sigma^2_{\mu})$$ 



Found marginal likelihood $\cal{L}(\mu, \sigma^2, \sigma^2_\mu)$ by integrating out $\mu_i$ with respect to $g$


$$\cal{L}(\mu, \sigma^2, \sigma^2_\mu) \propto \prod_{i = 1}^n (\sigma^2 + \sigma^2_\mu)^{-1/2} \exp \left\{ - \frac{1}{2} \frac{\left(y_i - \mu \right)^2}{\sigma^2 + \sigma^2_\mu }\right\}$$







Posterior for $\theta = \mu, \sigma^2_\mu$ with $\sigma^2 = 1$

$$\pi(\theta \mid y) = \frac{\pi(\theta) \cal{L}(\theta)}
{\int_\Theta \pi(\theta) \cal{L}(\theta) \, d\theta} =
\frac{\pi(\theta) \cal{L}(\theta)}
{m(y)}$$



---
##  Stochastic Sampling Intuition

- From a sampling perspective, we need to have a large sample or group of values, $\theta^{(1)}, \ldots, \theta^{(S)}$ from $\pi(\theta \mid  y)$ whose empirical distribution approximates $\pi(\theta \mid  y)$.

--

-  for any two sets $A$ and $B$, we want
.block[
.small[
$$\frac{\dfrac{\# \theta^{(s)} \in A}{S}}{\dfrac{\# \theta^{(s)} \in B}{S} } = \dfrac{\# \theta^{(s)} \in A}{\# \theta^{(s)} \in B} \approx \dfrac{\pi(\theta \in A \mid  y)}{\pi(\theta \in B \mid  y)}$$
]
]
--

- Suppose we have a working group $\theta^{(1)}, \ldots, \theta^{(s)}$ at iteration $s$, and need to add a new value $\theta^{(s+1)}$.

--

- Consider a candidate value $\theta^\star$ that is  _close_ to $\theta^{(s)}$.  


--

- Should we set $\theta^{(s+1)} = \theta^\star$ or not?

---
## Metropolis Ratio


look at  the ratio 




$$
\begin{split}
M & = \dfrac{\pi(\theta^\star \mid y)}{\pi(\theta^{(s)} \mid y)} = \frac{\dfrac{p(y \mid \theta^\star) \pi(\theta^\star)}{p(y)} } {\dfrac{p(y \mid \theta^{(s)}) \pi(\theta^{(s)})}{p(y)}}\\
\\
&  = \dfrac{p(y \mid \theta^\star) \pi(\theta^\star)}{p(y \mid \theta^{(s)}) \pi(\theta^{(s)})}
\end{split}
$$


--

- does not depend on the marginal likelihood we don't know!




---
## Metropolis algorithm

- If $M > 1$
  + Intuition: $\theta^{(s)}$ is already a part of the density we desire and the density at $\theta^\star$ is even higher than the density at $\theta^{(s)}$.
  
  + Action: set $\theta^{(s+1)} = \theta^\star$

--

- If $M < 1$, 
  + Intuition: relative frequency of values in our group $\theta^{(1)}, \ldots, \theta^{(s)}$ "equal" to $\theta^\star$ should be $\approx M = \dfrac{\pi(\theta^\star \mid y)}{\pi(\theta^{(s)} \mid y)}$. 
 
--
  + For every $\theta^{(s)}$, include only a fraction of an instance of $\theta^\star$.

--
  + Action: set $\theta^{(s+1)} = \theta^\star$ with probability $M$ and $\theta^{(s+1)} = \theta^{(s)}$ with probability $1-M$.



---
## Proposal Distribution


- Where should the proposed value $\theta^\star$ come from? 

--

- Sample $\theta^\star$ close to the current value $\theta^{(s)}$ using a **symmetric proposal distribution** $q(\theta^\star \mid \theta^{(s)})$

--

- $q()$ is actually a "family of proposal distributions", indexed by the specific value of $\theta^{(s)}$.

--

- Here, symmetric means that $q(\theta^\star \mid \theta^{(s)}) = q(\theta^{(s)} \mid \theta^\star)$. 

--

- Common choice 
$$\textsf{N}(\theta^\star; \theta^{(s)}, \delta^2 \Sigma)$$
 with $\Sigma$ based on the approximate $\textsf{Cov}(\theta \mid y)$ and $\delta = 2.44/\text{dim}(\theta)$ or tune
 
--

$$\text{Unif}(\theta^\star; \theta^{(s)} - \delta, \theta^{(s)} + \delta)$$


---
## Metropolis Algorithm Recap

- The algorithm proceeds as follows:

  1. Given $\theta^{(1)}, \ldots, \theta^{(s)}$, generate a _candidate_ value $\theta^\star \sim q(\theta^\star \mid \theta^{(s)})$.
  
--

  2. Compute the acceptance ratio
  .block[
.small[
$$
\begin{split}
M & = \dfrac{\pi(\theta^\star \mid y)}{\pi(\theta^{(s)} \mid y)} = \dfrac{p(y \mid \theta^\star) \pi(\theta^\star)}{p(y \mid \theta^{(s)}) \pi(\theta^{(s)})}.
\end{split}
$$
]
]

--

  3. Set
  .block[
.small[
\begin{eqnarray*}
\theta^{(s+1)} = \left\{ \begin{array}{ll}
\theta^\star & \quad \text{with probability} \quad \text{min}(M,1) \\
\theta^{(s)} & \quad \text{with probability} \quad 1 - \text{min}(M,1) \\
\end{array} \right.
\end{eqnarray*}
]
]

      equivalent to sampling $u \sim U(0,1)$ independently and setting
  .block[
.small[
\begin{eqnarray*}
\theta^{(s+1)} = \left\{ \begin{array}{ll}
\theta^\star & \quad \text{if} \quad u < M \\
\theta^{(s)} & \quad \text{if} \quad \text{otherwise} \\
\end{array} \right. .
\end{eqnarray*}
]
]


---
## Metropolis algorithm

- Once we obtain the samples, then we are back to using Monte Carlo approximations for quantities of interest!

--

- we can approximate posterior means, quantiles, and other quantities of interest using the empirical distribution of our sampled values.

--

- easy to compute the posterior distribution of nonlinear functions of parameters!

$$\psi^{(s)} = g(\theta^{(s)})$$

--

- some posterior summaries are hard to calculate based on samples $\{ \theta^{(s)}\}$ 

--
  + mode/MAP (at least for continuous)
  
--
  + marginal likelihood  $m(y) = \int \pi(\theta) p(y \mid \theta)\, d\theta$
---
##  Notes

-  The Metropolis chain ALWAYS moves to the proposed $\theta^\star$ at iteration $s+1$ if $\theta^\star$ has
higher target density than the current $\theta^{(s)}$.

--

- Sometimes, it also moves to a $\theta^\star$ value with lower density in proportion to the density value itself. 

--

- This leads to a random, Markov process that naturally explores the space according to the probability defined by $\pi(\theta \mid y)$, and hence generates a sequence that, while dependent, eventually represents draws from $\pi(\theta \mid y)$ (stationary distribution of the Markov Chain).




---
## Convergence

We will not cover the convergence theory behind Metropolis chains in detail, but ...

--

- The Markov process generated under this procedure is **ergodic** (irreducible and aperiodic) and has a unique limiting distribution (stationary distribution)

--

  + ergodicity means that the chain can move anywhere at each step, which is ensured, if $q(\theta^\star \mid \theta^{(s)}) > 0$ everywhere!

--

- By construction, Metropolis chains are **reversible**, so that  $\pi(\theta \mid y)$ is the stationary distribution
  
--

  + Think of reversibility as being equivalent to symmetry of the joint density of two consecutive $\theta^{(s)}$ and $\theta^{(s+1)}$ in the stationary process (which we get by using a symmetric proposal distribution)  
  
  + detailed balance
  




---
## Example

Priors with $\sigma^2 = 1$:

$$p(\mu) \propto 1$$
--

- Use a $\textsf{Cauchy}(0,1)$ prior on $\sigma_\mu$ independent of $\mu$ and 

--

- Symmetric proposal for $\mu$ and  $\sigma_\tau$?    

--

-  Try independent normals $\frac{2.44^2}{d} \textsf{Cov}(\theta)$ where $d$ is the dimension of $\theta$  (d = 2)



```{r metropolis, echo=FALSE}
tau = .1

sigma = 1
n = 100
mu.true = sort(rnorm(n, 0, tau))
y = rnorm(n, mu.true, sigma)

ybar = mean(y)
ssy = sum((y - ybar)^2)

R = 10000

#initialize

sigmamu = rep(NA, R)
mu = rep(NA, R)
sigmamu.prop = rep(NA, R)
mu.prop = mu
keep = mu

log.lik =  function(mu, sigma.mu, sigma, ybar, ssy, n ) {
  -.5*(n*log(sigma^2 + sigma.mu^2) + 
         (ssy +
         n*(ybar - mu)^2)/(sigma^2 + sigma.mu^2))
}

log.prior = function(x) { dt(x, 1, log=TRUE)}
sigmamu.sd = .1*2.44/2
mu.sd = .5*2.44/sqrt(n)

# initialize
sigmamu[1] = 1/tau
mu[1] = rnorm(1, ybar, sqrt(1 + sigmamu[1]^2)/sqrt(n))
log.post = log.prior(sigmamu[1]) + log.lik(mu[1],sigmamu[1], sigma, ybar, ssy, n)
keep[1] = 1
acc=0

for (r in 2:R) {
  sigmamu.star = rnorm(1, sigmamu[r - 1], sigmamu.sd)
  mu.star = rnorm(1, mu[r-1], mu.sd)
  log.prop = log.prior(sigmamu.star) + 
    log.lik(mu.star, sigmamu.star, sigma, ybar, ssy, n)
  M = min(1, exp(log.prop - log.post))
  sigmamu.prop[r] = sigmamu.star
  mu.prop[r] = mu.star
  if (runif(1) < M) {
    sigmamu[r] =  sigmamu.star
    mu[r] =  mu.star
    log.post = log.prop
    acc = acc + 1
    keep[r] = 1
  }
  else { 
         mu[r] = mu[r-1]
         sigmamu[r] = sigmamu[r-1]
         keep[r] = 2
  }
}
 
sigmamu = abs(sigmamu) 
theta1 = cbind(mu, sigmamu)
colnames(theta1) = c("mu", "sigma_mu")
```


```{r metropolis_2, echo=FALSE}




#initialize

sigmamu = rep(NA, R)
mu = rep(NA, R)
sigmamu.prop = rep(NA, R)
mu.prop = mu
keep = mu

log.lik =  function(mu, sigma.mu, sigma, ybar, ssy, n ) {
  -.5*(n*log(sigma^2 + sigma.mu^2) + 
         (ssy +
         n*(ybar - mu)^2)/(sigma^2 + sigma.mu^2))
}

log.prior = function(x) { dt(x, 1, log=TRUE)}


# initialize
sigmamu[1] = 1
mu[1] = rnorm(1, ybar, sqrt(1 + sigmamu[1]^2)/sqrt(n))
log.post = log.prior(sigmamu[1]) + log.lik(mu[1],sigmamu[1], sigma, ybar, ssy, n)
keep[1] = 1
acc=0

for (r in 2:R) {
  sigmamu.star = rnorm(1, sigmamu[r - 1], sigmamu.sd)
  mu.star = rnorm(1, mu[r-1], mu.sd)
  log.prop = log.prior(sigmamu.star) + 
    log.lik(mu.star, sigmamu.star, sigma, ybar, ssy, n)
  M = min(1, exp(log.prop - log.post))
  sigmamu.prop[r] = sigmamu.star
  mu.prop[r] = mu.star
  if (runif(1) < M) {
    sigmamu[r] =  sigmamu.star
    mu[r] =  mu.star
    log.post = log.prop
    acc = acc + 1
    keep[r] = 1
  }
  else { 
         mu[r] = mu[r-1]
         sigmamu[r] = sigmamu[r-1]
         keep[r] = 2
  }
}
 
sigmamu = abs(sigmamu) 
theta = cbind(mu, sigmamu)
colnames(theta) = c("mu", "sigma_mu")
```

---
##  First 200 Samples

```{r joint, fig=TRUE, echo = FALSE,out.width="75%", fig.height=3.5, fig.width=5}

use = 1:200
x = abs(sigmamu.prop[use])
y = mu[use]
plot(x, y, col= keep,  xlab=expression(sigma[mu]), ylab=expression(mu))

lines(sigmamu, mu, lty=2)
points(x, y, col= keep)
```


- Overall Acceptance probability is `r round(acc/(R - 1), 2)`   out of `r R` samples

--

- Goal is around 0.44 in 1 dimension to 0.23 in higher dimensions


---
## Tuning

- Sampled values are correlated

--

- Correlation between samples can be adjusted by selecting an optimal $\delta$ (i.e., spread of the distribution) in the proposal distribution

--

- $\delta$ too small leads to $M \approx 1$ for most proposed values, a high acceptance rate, but very small moves, leading to highly correlated chain.

--

- $\delta$ too large can get "stuck"  because $\theta^\star$ may be very far away from high density regions, leading to a very low acceptance rate and again high correlation in the Markov chain.

--


- Burn-in and thinning can help!



---
## Burn-in

- Convergence  occurs regardless of our starting point (in theory),  so we can usually pick any reasonable values in the parameter space as a starting point.

--

- May take a long time to reach high density regions

--
- Over representation of low density samples given finite iterations

--

- Generally, we throw out a certain number of the first draws, known as the **burn-in**, as an attempt to make our draws closer to the
stationary distribution and less dependent on any single set of starting values. 

--

- However, we don't know exactly when convergence occurs, so it is not always clear how much burn-in we would need.


--

- If you run long enough you should not need to discard any samples!  (ergodicity)







---
## Convergence diagnostics

- Diagnostics available to help decide on number of burn-in & collected samples.

--

- **Note**: no definitive tests of convergence but you should do as many diagnostics as you can, on all parameters in your model.

--

- With "experience", visual inspection of trace plots perhaps most useful approach.

--

- There are a number of useful automated tests in R.

--

- **CAUTION**: diagnostics cannot guarantee that a chain has converged, but they can indicate it has not converged.



---
## Diagnostics in R




- The most popular package for MCMC diagnostics in R is `coda`.

--

- `coda` uses a special MCMC format so you must always convert your posterior matrix into an MCMC object. 

--

- For the example, we have the following in R.
  ```{r fig.height=4}
#library(coda)
theta.mcmc <- mcmc(theta,start=1) #no burn-in (simple problem!)
```



---
## Diagnostics in R

```{r fig.height=4}
summary(theta.mcmc)
```

The naive SE is the **standard error of the mean**, which captures simulation error of the mean rather than the posterior uncertainty. 

The time-series SE adjusts the naive SE for **autocorrelation**.



---
## Effective sample size

- The **effective sample size** translates the number of MCMC samples $S$ into an equivalent number of independent samples.

--

- It is defined as
.block[
.small[
$$\textrm{ESS} = \dfrac{S}{1 + 2 \sum_k \rho_k},$$
]
]

  where $S$ is the sample size and $\rho_k$ is the lag $k$ autocorrelation.
  
--

- For our data, we have
  ```{r }
effectiveSize(theta.mcmc)
```

```{r, echo=FALSE,include=FALSE}
ESS = effectiveSize(theta.mcmc)
```

--

- So our 10,000 samples are equivalent to `r round(ESS["mu"], 1)` independent samples for $\mu$  and `r round(ESS["sigma_mu"], 1)` independent samples for $\sigma_\mu$.




---
## Trace plot for mean

```{r fig.height=3.9}
plot(theta.mcmc[,"mu"])
```



---
## Trace plot for $\sigma_\mu$

```{r fig.height=3.9}
plot(theta.mcmc[,"sigma_mu"])
```

OK  (be careful of scaling in plots!)



---
## Autocorrelation

- Another way to evaluate convergence is to look at the autocorrelation between draws of our Markov chain.

--

- The lag $k$ autocorrelation, $\rho_k$, is the correlation between each draw and its $k$th lag, defined as
.block[
.small[
$$\rho_k = \dfrac{\sum_{s=1}^{S-k}(\theta_s - \bar{\theta})(\theta_{s+k} - \bar{\theta})}{\sum_{s=1}^{S-k}(\theta_s - \bar{\theta})^2}.$$
]
]

--

- We expect the autocorrelation to decrease as $k$ increases. 

--

- If autocorrelation remains high as $k$ increases, we have slow mixing due to the inability of the sampler to move around the space well.



---
## Autocorrelation for mean

```{r fig.height=3.9}
autocorr.plot(theta.mcmc[,"mu"])
```

So-So


---
## Autocorrelation for variance 

```{r fig.height=3.9}
autocorr.plot(theta.mcmc[,"sigma_mu"])
```

worse




---
## Gelman-Rubin 

Gelman  & Rubin suggested a diagnostic $R$ based on taking separate  chains with dispersed initial values to test convergence

--
```{r, fig.height=4, echo=FALSE}
plot(theta1[1:500,"sigma_mu"], col=2, ylab=expression(sigma[mu]))
points(1:500, theta[1:500, "sigma_mu"], col=1)
```


---
## Gelman-Rubin Diagnostic

- Run m > 2 chains of length 2S from overdispersed starting values.
- Discard the first S draws in each chain.
-  Calculate the pooled within-chain variance $W$ and between-chain variance $B$.

--

.block[
.small[
$$R = \frac{\frac{S-1}{S} W + \frac{1}{S} B }{W}$$
]
]

--

- numerator and denominator are both unbiased estimates of the variance if the two chains have converged

--
  +  otherwise $W$ is an underestimate (hasn't explored enough)
  +  numerator will overestimate as $B$ is too large (overdispersed starting points)
  
--

- As $S \to \infty$ and $B \to 0$,  $R \to 1$

--

- version in `R` is slightly different 

---
## Gelman-Rubin Diagnostic

```{r, echo=FALSE}
theta2 = theta
```

```{r}
theta.mcmc = mcmc.list(mcmc(theta1, start=5000), mcmc(theta2, start=5000))
gelman.diag(theta.mcmc)
```

--

-  Values of $R > 1.1$ suggest lack of convergence

--
- Looks OK
 
--

See also `gelman.plot`

---
## Geweke statistic

- Geweke proposed taking two non-overlapping parts of a single Markov chain (usually the first 10% and the last 50%) and comparing the mean of both parts, using a difference of means test

--

- The null hypothesis would be that the two parts of the chain are from the same distribution. 

--

- The test statistic is a z-score with standard errors adjusted for autocorrelation, and if the p-value is significant for a variable, you need more draws. 

---
## Geweke Diagnostic

- The output is the z-score itself (not the p-value).

```{r fig.height=3.5}
geweke.diag(theta.mcmc)
```


---
## Practical advice on diagnostics

- There are more tests we can use: Raftery and Lewis diagnostic, Heidelberger and Welch, etc.

--

- The Gelman-Rubin approach is quite appealing in using multiple chains

--

- Geweke (and Heidelberger and Welch) sometimes reject even when the trace plots look good.

--

- Overly sensitive to minor departures from stationarity that do not impact inferences.

--



- Most common method of assessing convergence is visual examination of trace plots.


---
##  Improving

-  more iterations and multiple chains

--

- thinning to reduce correlations and increase ESS

--

- change the proposal distribution $q$




