---
title: "Lecture 9: More MCMC: Adaptive Metropolis, Metropolis-Hastings,  and Gibbs"
author: "Merlise Clyde"
date: "September 23"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(lattice)
require(MCMCpack)
require(hdrcde)
require(coda)
require(mvtnorm)
#require(rethinking)
set.seed(42)
```




## Example from Last Class

Priors with $\sigma^2 = 1$:

$$p(\mu) \propto 1$$
--

- Use a $\textsf{Cauchy}(0,1)$ prior on $\sigma_\mu$ independent of $\mu$ and 

--

- Symmetric proposal for $\mu$ and  $\sigma_\tau$    

--

-  Independent normals centered at current values of $\mu$ and $\sigma_\mu$  with covariance $\frac{2.4^2}{d} \textsf{Cov}(\theta)$ where $d = 2$ (the dimension of $\theta$ ) 



```{r metropolis, echo=FALSE}
tau = .1

sigma = 1
n = 100
mu.true = sort(rnorm(n, 0, tau))
y = rnorm(n, mu.true, sigma)

ybar = mean(y)
ssy = sum((y - ybar)^2)

R = 10000

#initialize

sigmamu = rep(NA, R)
mu = rep(NA, R)
sigmamu.prop = rep(NA, R)
mu.prop = mu
keep = mu

log.lik =  function(mu, sigma.mu, sigma, ybar, ssy, n ) {
  -.5*(n*log(sigma^2 + sigma.mu^2) + 
         (ssy +
         n*(ybar - mu)^2)/(sigma^2 + sigma.mu^2))
}

log.prior = function(x) { dt(x, 1, log=TRUE)}
sigmamu.sd = .1*2.44/2
mu.sd = .5*2.44/sqrt(n)

# initialize
sigmamu[1] = 1/tau
mu[1] = rnorm(1, ybar, sqrt(1 + sigmamu[1]^2)/sqrt(n))
log.post = log.prior(sigmamu[1]) + log.lik(mu[1],sigmamu[1], sigma, ybar, ssy, n)
keep[1] = 1
acc=0

for (r in 2:R) {
  sigmamu.star = rnorm(1, sigmamu[r - 1], sigmamu.sd)
  mu.star = rnorm(1, mu[r-1], mu.sd)
  log.prop = log.prior(sigmamu.star) + 
    log.lik(mu.star, sigmamu.star, sigma, ybar, ssy, n)
  M = min(1, exp(log.prop - log.post))
  sigmamu.prop[r] = sigmamu.star
  mu.prop[r] = mu.star
  if (runif(1) < M) {
    sigmamu[r] =  sigmamu.star
    mu[r] =  mu.star
    log.post = log.prop
    acc = acc + 1
    keep[r] = 1
  }
  else { 
         mu[r] = mu[r-1]
         sigmamu[r] = sigmamu[r-1]
         keep[r] = 2
  }
}
 
sigmamu = abs(sigmamu) 
theta1 = cbind(mu, sigmamu)
colnames(theta1) = c("mu", "sigma_mu")
```


```{r metropolis_2, echo=FALSE}




#initialize

sigmamu = rep(NA, R)
mu = rep(NA, R)
sigmamu.prop = rep(NA, R)
mu.prop = mu
keep = mu

log.lik =  function(mu, sigma.mu, sigma, ybar, ssy, n ) {
  -.5*(n*log(sigma^2 + sigma.mu^2) + 
         (ssy +
         n*(ybar - mu)^2)/(sigma^2 + sigma.mu^2))
}

log.prior = function(x) { dt(x, 1, log=TRUE)}


# initialize
sigmamu[1] = 1
mu[1] = rnorm(1, ybar, sqrt(1 + sigmamu[1]^2)/sqrt(n))
log.post = log.prior(sigmamu[1]) + log.lik(mu[1],sigmamu[1], sigma, ybar, ssy, n)
keep[1] = 1
acc=0

for (r in 2:R) {
  sigmamu.star = rnorm(1, sigmamu[r - 1], sigmamu.sd)
  mu.star = rnorm(1, mu[r-1], mu.sd)
  log.prop = log.prior(sigmamu.star) + 
    log.lik(mu.star, sigmamu.star, sigma, ybar, ssy, n)
  M = min(1, exp(log.prop - log.post))
  sigmamu.prop[r] = sigmamu.star
  mu.prop[r] = mu.star
  if (runif(1) < M) {
    sigmamu[r] =  sigmamu.star
    mu[r] =  mu.star
    log.post = log.prop
    acc = acc + 1
    keep[r] = 1
  }
  else { 
         mu[r] = mu[r-1]
         sigmamu[r] = sigmamu[r-1]
         keep[r] = 2
  }
}
 
sigmamu = abs(sigmamu) 
theta = cbind(mu, sigmamu)
colnames(theta) = c("mu", "sigma_mu")
```


---
## Gelman-Rubin 

Gelman  & Rubin suggested a diagnostic $R$ based on taking separate  chains with dispersed initial values to test convergence

--
```{r, fig.height=4, echo=FALSE}
plot(theta1[1:500,"sigma_mu"], col=2, ylab=expression(sigma[mu]))
points(1:500, theta[1:500, "sigma_mu"], col=1)
```


---
## Gelman-Rubin Diagnostic

- Run m > 2 chains of length 2S from overdispersed starting values.
- Discard the first S draws in each chain.
-  Calculate the pooled within-chain variance $W$ and between-chain variance $B$.

--

.block[
.small[
$$R = \frac{\frac{S-1}{S} W + \frac{1}{S} B }{W}$$
]
]

--

- numerator and denominator are both unbiased estimates of the variance if the two chains have converged

--

  +  otherwise $W$ is an underestimate (hasn't explored enough)
  
--
  +  numerator will overestimate as $B$ is too large (overdispersed starting points)
  
--

- As $S \to \infty$ and $B \to 0$,  $R \to 1$

--

- Note: version in `R` is slightly different 

---
## Gelman-Rubin Diagnostic

```{r, echo=FALSE}
theta2 = theta
```

```{r}
theta.mcmc = mcmc.list(mcmc(theta1, start=5000), mcmc(theta2, start=5000))
gelman.diag(theta.mcmc)
```

--

-  Values of $R > 1.1$ suggest lack of convergence

--
- Looks OK
 
--

See also `gelman.plot`

---
## Geweke statistic

- Geweke proposed taking two non-overlapping parts of a single Markov chain (usually the first 10% and the last 50%) and comparing the mean of both parts, using a difference of means test

--

- The null hypothesis would be that the two parts of the chain are from the same distribution. 

--

- The test statistic is a z-score with standard errors adjusted for autocorrelation, and if the p-value is significant for a variable, you need more draws. 

---
## Geweke Diagnostic

- The output is the z-score itself (not the p-value).

```{r fig.height=3.5}
geweke.diag(theta.mcmc)
```


---
## Practical advice on diagnostics

- There are more tests we can use: Raftery and Lewis diagnostic, Heidelberger and Welch, etc.

--

- The Gelman-Rubin approach is quite appealing in using multiple chains

--

- Geweke (and Heidelberger and Welch) sometimes reject even when the trace plots look good.

--

- Overly sensitive to minor departures from stationarity that do not impact inferences.

--



- Most common method of assessing convergence is visual examination of trace plots.


---
##  Improving

-  more iterations and multiple chains

--

- thinning to reduce correlations and increase ESS

--

- change the proposal distribution $q$



---
## Proposal Distribution




Common choice 
$$\textsf{N}(\theta^\star; \theta^{(s)}, \delta^2 \Sigma)$$
--

-  rough estimate of $\Sigma$ based on the asymptotic Gaussian approximation $\textsf{Cov}(\theta \mid y)$ and $\delta = 2.38/\sqrt{\text{dim}(\theta)}$ 

--
  + find the MAP estimate (posterior mode)  $\hat{\theta}$
  
--
  + take 
  $$\Sigma =  \left[- 
  \frac{\partial^2 \log(\cal{L}(\theta)) + \log(\pi(\theta))}
       {\partial \theta \partial \theta^T} \right]^{-1}_{\theta = \hat{\theta}}$$
  
--

- ignore prior and use inverse of Fisher Information (covariance of MLE)
---
### Adaptive Metropolis?

-  MCMC doesn't allow you to use the full history of the chain $\theta^{(1)}, \ldots, \theta^{(s)}$ in constructing the proposal distributions 

--

- violates the Markov assumption

--

- Workaround?  run an initial MCMC for an initial tuning phase (e.g. 1000 samples) and then fix the kernel to depend only on $\theta^{(s-1)}$ and $y$.


--

- more elegant approach -  formal **adaptive Metropolis**

--

  + keep adapting the entire time!
  
--
  + this may mess up convergence !  
  
--
  + need conditions for vanishing adaptation e.g.  that the proposal depends less and less on recent states in the chain - Roberts & Rosenthal (2006) and other conditions
  
  
---
## Adaptive MCMC

- Haario et al (2001) propose a simple and effective adaptive random walk Metropolis  (RWM)

--

-  run  RWM with a Gaussian proposal for a fixed number of iterations for $s < s_0$

--

- estimate of covariance at state $s$

$$\Sigma^{(s)} = \frac{1}{s}\left(\sum_{i=1}^s \theta^{(i)} {\theta^{(i)}}^T - 
s \bar{\theta}^{(s)} {\bar{\theta}^{(s)}}^T\right)$$

--

- proposal for $s > s_0$ with $\delta = 2.38/\sqrt{d}$

$$\theta^* \sim \textsf{N}(\theta^{(s)}, \delta^2 (\Sigma^{(s)} + \epsilon I_d))$$

--

- $\epsilon > 0$ insures covariance is positive definite

--

- if $s_0$ is too large will take longer for adaptation to be seen

---
## Example again
```{r adaptive_metropolis, echo=FALSE}
require(mvtnorm)
tau = .10
eps = .01

sigma = 1
n = 100
mu.true = sort(rnorm(n, 0, tau))
y = rnorm(n, mu.true, sigma)

adapt.MH = function(y, R=10000, B=200, sigma.0 = 1) {
  ybar = mean(y)
  ssy = sum((y - ybar)^2)

#initialize

  theta = matrix(NA, ncol=2, nrow=R)

  log.lik =  function(theta, ybar, ssy, n ) {
    sigma = 1
    mu = theta[1]
    sigma.mu = theta[2]
    -.5*(n*log(sigma^2 + sigma.mu^2) + 
         (ssy + n*(ybar - mu)^2)/(sigma^2 + sigma.mu^2))
}

log.prior = function(theta) { dt(theta[2], 1, log=TRUE)}

Sigma = diag(c(1/n, sigma.0^2))*2.38^2/2

# initialize
theta[1,] = c(ybar, sigma.0)


log.post = log.prior(theta[1,]) + log.lik(theta[1,], ybar, ssy, n)
acc=0

# burnin
for (r in 2:B) {
  theta.star = rmvnorm(1, theta[r-1,], Sigma)
  log.prop = log.prior(theta.star) + log.lik(theta.star, ybar, ssy, n)
  M = min(1, exp(log.prop - log.post))
  if (runif(1) < M) {
    theta[r,] = theta.star
    log.post = log.prop
    acc = acc + 1
  }
  else { 
         theta[r,] = theta[r-1,]
  }
}

 theta.cov = (2.38^2/2)*(cov(theta[1:B,]) + eps*diag(rep(1,2)))
 eps = min(diag(theta.cov))/100
 
 #start adaptive sample
 acc = 0
 for (r in (B+1):R) {
  theta.star = rmvnorm(1, theta[r-1,], theta.cov)
  log.prop = log.prior(theta.star) + log.lik(theta.star, ybar, ssy, n)
  M = min(1, exp(log.prop - log.post))
  if (runif(1) < M) {
    theta[r,] = theta.star
    log.post = log.prop
    acc = acc + 1
  }
  else { 
         theta[r,] = theta[r-1,]
  }
   theta.cov = (2.38^2/2)*(cov(theta[1:r,]) + eps*diag(rep(1,2)))

}
 
 acc = acc/(R - B)
 colnames(theta) = c("mu", "sigma_mu")
 theta[,2] = abs(theta[,2])
 theta = mcmc(theta)
 return(list(theta=theta, acc=acc))
}

out1 = adapt.MH(y, sigma.0=1)

out2 = adapt.MH(y, sigma.0 = 5)

out = mcmc.list(out1$theta, out2$theta)
```

```{r, fig.height=4, echo=FALSE}
show = 1:2000
plot(as.numeric(out2$theta[show,2]), type="l", ylab=expression(sigma[mu]), xlab="state s")
abline(v=200)
lines(as.numeric(out1$theta[show,2]), col=2)
```

Acceptance rate now around 30-35 % of 10,000 iterations!
---
##  Metropolis-Hastings  (MH)

- Metropolis requires that the proposal distribution be symmetric

--

- Hastings (1970)  generalizes Metropolis algorithms to allow asymmetric proposals - aka Metropolis-Hastings or MH  $q(\theta^* \mid \theta^{(s)})$ does not need to be the same as $q(\theta^{(s)} \mid \theta^*)$

--
-  propose $\theta^*  \mid \theta^{(s)} \sim q(\theta^* \mid \theta^{(s)})$

--
- Acceptance probability 

$$\min \left\{ 1, \frac{\pi(\theta^*) \cal{L}(\theta^*)/q(\theta^* \mid \theta^{(s)})}
{\pi(\theta^{(s)}) \cal{L}(\theta^{(s)})/q( \theta^{(s)} \mid \theta^*)} \right\}$$

--

- adjustment for asymmetry in acceptance ratio is key to ensuring convergence to stationary distribution!

---
## Special cases

- Metropolis

--

- Independence chain

--

- Gibbs samplers

--

- Metropolis-within-Gibbs

--

- combinations of the above!

---
## Independence Chain

- suppose we have a good approximation $\tilde{\pi}(\theta \mid y)$   to  $\pi(\theta \mid y)$

--

-  Draw $\theta^* \sim \tilde{\pi}(\theta \mid y)$ _without_ conditioning on $\theta^{(s)}$

--

- acceptance probability 

$$\min \left\{ 1, \frac{\pi(\theta^*) \cal{L}(\theta^*)/\tilde{\pi}(\theta^* \mid \theta^{(s)})}
{\pi(\theta^{(s)}) \cal{L}(\theta^{(s)})/\tilde{\pi}( \theta^{(s)} \mid \theta^*)} \right\}$$

--

- what happens if the approximation is really accurate?

--

- probability of acceptance is $\approx 1$

--

- Important caveat for convergence:  tails of the posterior should be at least as heavy as the tails of the posterior  (Tweedie 1994)

--

-  Replace Gaussian by a Student-t with low degrees of freedom

--

- transformations of $\theta$

---
## Block Updates & Gibbs

So far all algorithms update all of the parameters simultaneously

--

- convenient to break problems in to $K$ blocks and update them separately

--

- $\theta = (\theta_{[1]}, \ldots, \theta_{[K]}) = (\theta_1, \ldots, \theta_p)$



--

At iteration $s$, for $k = 1, \ldots, K$ Cycle thru blocks: (fixed order or random order)
   + propose $\theta^*_{[k]} \sim q_k(\theta_{[k]} \mid \theta_{[<k]}^{(s)}, \theta_{[>k]}^{(s-1)})$
   
   + set $\theta_{[k]}^{(s)} = \theta^*_{[k]}$ with probability
   
$$\min \left\{ 1, \frac{
   \pi(\theta_{[<k]}^{(s)},\theta_{[k]}^*, 
        \theta_{[>k]}^{(s-1)})
   \cal{L}(\theta_{[<k]}^{(s)},\theta_{[k]}^*,
           \theta_{[>k]}^{(s-1)})/
   q_k(\theta_{[k]}^* \mid \theta_{[<k]}^{(s)},    
       \theta_{[>k]}^{(s-1)})}
  {\pi(\theta_{[<k]}^{(s)},\theta_{[k]}^{(s-1)}, 
        \theta_{[>k]}^{(s-1)})
   \cal{L}(\theta_{[<k]}^{(s)},\theta_{[k]}^{(s-1)},
           \theta_{[>k]}^{(s-1)})/
   q_k(\theta_{[k]}^{(s-1)} \mid \theta_{[<k]}^{(s)},    
       \theta_{[>k]}^{(s-1)})} \right\}$$
---
## Gibbs Sampler

special case of MH

--

- proposal distribution $q_k$ for the $k$th block is the **full conditional** distribution for $\theta_{[k]}$

$$\pi(\theta_{[k]} \mid \theta_{[-k]}, y) = \frac{\pi(\theta_{[k]} , \theta_{[-k]} \mid y)}{ \pi(\theta_{[-k]} \mid y))} \propto \pi(\theta_{[k]} , \theta_{[-k]} \mid y)$$
--

$$\pi(\theta_{[k]} \mid \theta_{[-k]}, y)  \propto \cal{L}(\theta_{[k]} , \theta_{[-k]})\pi(\theta_{[k]} , \theta_{[-k]})$$
--

- acceptance probability is always 1!

--

- even though joint distribution is messy, full conditionals may be (conditionally) conjugate and easy to sample from!

---
## Comments

- can use Gibbs steps and Metropolis Hastings steps together

--

- Use block sizes in Gibbs that are as big as possible to improve  mixing (proven faster convergence)

--

- combine with adaptive Metropolis 

--

- Adaptive Independence Metropolis Hastings  (learn a mixture)