---
title: "Lecture 11: More Gibbs:  Regression and Data Augmentation"
author: "Merlise Clyde"
date: "September 30"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(lattice)
require(MCMCpack)
require(hdrcde)
require(coda)
require(mvtnorm)
#require(rethinking)
set.seed(42)
```


## Normal Linear Regression Example

Model
$$\begin{align*}
Y \mid \beta, \phi & \sim \textsf{N}(X \beta, \phi^{-1} I_n) \\
\beta & \sim \textsf{N}(b_0, \Phi_0^{-1}) \\
\phi & \sim \textsf{Gamma}(v_0/2, s_0/2)
\end{align*}$$



- Full Conditional for $\beta$

$$\begin{align*}
\beta & \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(b_n, \Phi_n^{-1}) \\
b_n & =  (\Phi_0 + \phi X^TX)^{-1}(\Phi_0 b_0  +  \phi X^TX \hat{\beta})\\
\Phi_n & = \Phi_0 + \phi X^TX
\end{align*}$$

- Full Conditional for $\phi$

$$ \phi \mid \beta, y_1, \ldots, y_n \sim \textsf{Gamma}((v_0 + n)/2, (s_0 + \sum_i(y_i - x^T_i \beta)))$$

- Choice of Prior Precision $\Phi_0$

---
## Invariance and Choice of Mean/Precision


- What if we transform the $X$ matrix by $\tilde{X} = X H$ where $H$ is $p \times p$ and invertible

--

- obtain the posterior for $\tilde{\beta}$ using $Y$ and $\tilde{X}$  

$$ Y \sim \textsf{N}_n (\tilde{X}\tilde{\beta}, \phi^{-1} I_n)$$

- invariance suggests that since $\tilde{X} \tilde{\beta} = X H  \tilde{\beta} = X \beta$  the posterior for $\beta$ and $H \tilde{\beta}$ should be the same 
(or the posterior of $H^{-1} \beta$
and $\tilde{\beta}$ should be the same)
--

- with some linear algebra we can show that this is true if $b_0 = 0$ and $\Phi_0$ is $k X^TX$ for some $k$  (show!)

---
## Zellner's g-prior

Popular choice is to take $k = \phi/g$ which is a special case of Zellner's g-prior

$$\beta \mid \phi, g \sim \textsf{N}\left(0, \frac{g}{\phi} (X^TX)^{-1}\right)$$

--

- Full conditional 

$$\beta \mid \phi, g \sim \textsf{N}\left(\frac{g}{1 + g} \hat{\beta}, \frac{1}{\phi} \frac{g}{1 + g} (X^TX)^{-1}\right)$$
--

- one parameter $g$ controls shrinkage

--

if $\phi \sim \textsf{Gamma}(v_0/2, s_0/2)$ then posterior is
$$\phi \mid y_1, \ldots, y_n \sim \textsf{Gamma}(v_n/2, s_n/2)$$
--

Conjugate so we could skip Gibbs sampling and sample directly from gamma and then conditional normal!

---
## Ridge Regression 

If $X^TX$ is nearly singular, certain  elements of $\beta$ or (linear combinations of $\beta$) may have huge variances under the $g$-prior (or flat prior) as the MLEs are highly unstable!

--

**Ridge regression** protects against the explosion of variances and ill-conditioning with the conjugate priors:


$$\beta \mid \phi \sim \textsf{N}(0, \frac{1}{\phi \lambda} I_p)$$
--

Posterior for $\beta$  (conjugate case)

$$\beta \mid \phi, \lambda, y_1, \ldots, y_n \sim 
\textsf{N}\left((\lambda I_p + X^TX)^{-1} X^T Y,  \frac{1}{\phi}(\lambda I_p + X^TX)^{-1}
\right)$$




---
##  Bayes Regression

- Posterior mean (or mode) given $\lambda$ is biased, but can show that there **always** is a value of $\lambda$  where the frequentist's expected squared error loss is smaller for the Ridge estimator than MLE!

--

- related to penalized maximum likelihood estimation 

--

-  Choice of $\lambda$

--

-  usual center and standardized $X$ !

--

-  Bayes Regression and choice of $\Phi_0$ in general is a very important problem and provides the foundation  for many variations on shrinkage estimators, variable selection, hierarchical models, nonparameteric regression and more!

--

- Be sure that you can derive the full conditional posteriors for $\beta$ and $\phi$ as well as the joint posterior in the conjugate case!


---
### Binary Regression

$$Y_i \mid \beta \sim \textsf{Ber}(p(x_i^T \beta))$$
where $\Pr(Y_i  = 1 \mid \beta) = p(x_i^T \beta))$ and linear predictor $x_i^T\beta = \lambda_i$

--

- link function for binary regression is any 1-1 function  $g$ that will map $(0,1) \to \mathbb{R}$,  i.e.  $g(p(\lambda)) = \lambda$

--

- logistic regression use the logit link  
$$\log\left(\frac{p(\lambda_i)}{1 - p(\lambda_i) }\right) = x_i^T \beta = \lambda_i$$
--

- probit link

$$p(x_i^T \beta) = \Phi(x_i^T \beta)$$
- $\Phi()$ is the Normal cdf

---
##  Likelihood and Posterior

Likelihood:

$$\cal{L}(\beta) \propto \prod_{i = 1}^n \Phi(x_i^T \beta)^{y_i} (1 - \Phi(x^T_i \beta))^{1 - y_i}$$
--

- prior  $\beta \sim \textsf{N}_p(b_0, \Phi_0)$

--

- posterior $\pi(\beta) \propto \pi(\beta) \cal{L}(\beta)$

--

- How to do approximate the posterior?

--
  + asymptotic Normal approximation
  
--
  + MH or adaptive Metropolis
  
--
  + stan (Hamiltonian Monte Carlo)
  
--
  + Gibbs ?   
  
--

seemingly no, but there is a trick!
  
  
---
##  Data Augmentation

-  Consider an **augmented** posterior

   $$\pi(\beta, Z \mid y) \propto \pi(\beta) \pi(Z \mid \beta) \pi(y \mid Z, \theta)$$
--

- IF we choose $\pi(Z \mid \beta) \pi(y \mid Z, \theta)$ carefully, we can carry out Gibbs and get samples of $\pi(\beta \mid y)$ 

--

Conditions: we need to have 

$$\pi(\beta \mid y) = \int_{\cal{Z}} \pi(\beta, z \mid y) \, dz$$
(it is a marginal of joint augmented posterior)

--

-  We have to choose

$$p(y \mid \theta) = \int_{\cal{Z}}  \pi(z \mid \beta) \pi(y \mid \beta, z) \, dz$$
--

- complete data likelihood

---
##  Augmentation Strategy

Set 

- $y_i = 1(Z_i > 0)$ i.e. ( $y_i = 1$ if $Z_i > 0$ ) 
- $y_i = 1(Z_i < 0)$ i.e.  ( $y_i = 0$ if $Z_i < 0$ )

--

- $Z_i = x_i^T \beta + \epsilon_i \qquad \epsilon_i \overset{iid}{\sim} \textsf{N(0,1)}$

--

- Relationship to probit model:

$$\begin{align*}\Pr(y = 1 \mid \beta) & = P(Z_i > 0 \mid \beta) \\
   & = P(Z_i - x_i^T \beta > -x^T\beta) \\
   & = P(\epsilon_i > -x^T\beta) \\
   & =  1 - \Phi(-x^T_i \beta) \\
   & =  \Phi(x^T_i \beta)
\end{align*}$$




---
##  Augmented Posterior & Gibbs

$$\begin{align*}\pi(& Z_1,  \ldots, Z_n,  \, \beta \mid y) \propto \\
& \textsf{N}(\beta; b_0, \phi_0)  \left\{\prod_{i=1}^n \textsf{N}(Z_i; x_i^T\beta, 1)\right\} \left\{  \prod_{i=1}^n y_i 1(Z_i > 0) + (1 - y_i)1(Z_i < 0)\right\}
\end{align*}$$

--
- full conditional for $\beta$
$$\beta \mid Z_1, \ldots, Z_n, y_1, \ldots, y_n \sim \textsf{N}(b_n, \Phi_n)$$   
- standard Normal-Normal regression updating given $Z_i$'s  

--
- Full conditional for latent $Z_i$

$$\begin{align*}
\pi(Z_i \mid \beta, Z_{[-i]}, y_1, \ldots, y_n)  & \propto 
\textsf{N}(Z_i; x_i^T \beta, 1)1(Z_i > 0) \text{   if  } y_1 = 1 \\
\pi(Z_i \mid \beta, Z_{[-i]}, y_1, \ldots, y_n)  & \propto 
\textsf{N}(Z_i; x_i^T \beta, 1)1(Z_i < 0) \text{   if  } y_1 = 0 \\
\end{align*}$$

- sample from independent truncated normal distributions !

--

- two block Gibbs sampler $\theta_{[1]} = \beta$ and $\theta_{[2]} = (Z_1, \ldots, Z_n)^T$

---
## Truncated Normal Sampling

- Use inverse cdf method for  cdf $F$

- If $u \sim U(0,1)$ set $z = F^{-1}(u)$

```{r inverseCDF, echo=FALSE, fig.height=5, out.height="55%", out.width="60%"}
a = pnorm(0); b = 1
Z = seq(-3.5, 3.5, length=1000) + .5
plot(Z, pnorm(Z), type="l", ylab="u")
u = .75
z = qnorm(u) 
segments(min(Z), u,  z, u, lty=2)
segments(z,0, z, u, lty=2)

abline(h = a, lty=3)
abline(h=b, lty=3)
```
- if $Z \in (a, b)$, Draw $u \sim U(F(a),F(b))$
  and set $z = F^{-1}(u)$

---
## Data Augmentation in General

DA is a broader than a computational trick allowing Gibbs sampling

--

- missing data

--

- random effects or latent variable modeling i.e we introduce latent variables to simplify dependence structure modelling


--

- Modeling heavy tailed distributions such as $t$ errors in regression 











---
## Comments



- Why don't we treat each individual $\beta_j$ as a separate block?

--

-  Gibbs always accepts, but can mix slowly if parameters in different blocks are highly correlated!

--

- Use block sizes in Gibbs that are as big as possible to improve  mixing (proven faster convergence)

--

- Collapse the sampler by integrating out as many parameters as possible  (as long as resulting sampler has good mixing)

--

- can use Gibbs steps and (adaptive) Metropolis Hastings steps together

--


- latent variables to allow Gibbs steps  but not always better!





