---
title: "STA 601: Bayesian Model Choice in Linear Regression"
subtitle: "STA 601 Fall 2021"
author: "Merlise Clyde"
date: "October 19, 2021"
header-includes:
  - \include{macros.tex}
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Bayesian Model Choice

**General setting:**

1. Define a list of models; let $\Gamma$ be a "finite" set of different possible models. 
  
--

2. Each model $\gamma$ is in $\Gamma$, including the "true" model. Also, let $\theta_\gamma$ represent the parameters in model $\gamma$.
  
--

3. Put a prior over the set $\Gamma$. Let $\Pi_\gamma = p[\gamma] = \Pr[\gamma \text{ is true}]$, for all $\gamma \in \Gamma$.
  
--

4. Put a prior on the parameters in each model, that is, each $\pi(\theta_\gamma)$.

--

5. Compute marginal posterior probabilities $\Pr[\gamma | Y]$ for each model, and select a model based on the posterior probabilities or use the full posterior over all models!

---
## Bayesian Model Probabilities

- For each model $\gamma \in \Gamma$, we need to compute $\Pr[\gamma | Y]$. 

--

- Let $p_\gamma(Y)$ denote the marginal likelihood of the data under model $\gamma$, that is, $\mathcal{p}[Y | \gamma]$. As before,
.block[
.small[
$$
\begin{split}
\hat{\Pi}_\gamma = \Pr[\gamma | Y] & = \frac{\mathcal{p}[Y | \gamma] \cdot p[\gamma]}{\sum_{\gamma^\star \in \Gamma} \mathcal{p}[Y | \gamma^\star] \cdot p[\gamma^\star]} = \frac{ p_\gamma(Y) \Pi_\gamma }{ \sum_{\gamma^\star \in \Gamma} p_{\gamma^\star}(Y) \Pi_{\gamma^\star} }\\
\\
& = \frac{ \Pi_\gamma \cdot \left[ \int_{\Theta_\gamma} p_\gamma(Y | \theta_\gamma) \cdot \pi(\theta_\gamma) \text{d}\theta_\gamma \right] }{ \sum_{\gamma^\star \in \Gamma} \Pi_{\gamma^\star} \cdot \left[ \int_{\Theta_{\gamma^\star}} p_{\gamma^\star}(Y | \theta_{\gamma^\star}) \cdot \pi(\theta_{\gamma^\star}) \text{d}\theta_{\gamma^\star} \right] }.\\
\end{split}
$$
]
]

--

- If we assume a uniform prior on $\Gamma$, that is, $\Pi_\gamma = \frac{1}{\#\Gamma}$, for all $\gamma \in \Gamma$, then
.block[
.small[
$$\hat{\Pi}_\gamma  = \frac{ p_\gamma(Y) }{ \sum_{\gamma^\star \in \Gamma} p_{\gamma^\star}(Y)  }
= \frac{ \left[ \int_{\Theta_\gamma} p_\gamma(Y | \theta_\gamma) \cdot \pi(\theta_\gamma) \text{d}\theta_\gamma \right] }{ \sum_{\gamma^\star \in \Gamma} \left[ \int_{\Theta_{\gamma^\star}} p_{\gamma^\star}(Y | \theta_{\gamma^\star}) \cdot \pi(\theta_{\gamma^\star}) \text{d}\theta_{\gamma^\star} \right]  }$$
]]


---
## Bayesian Model Selection 

- <div class="question">
How should we choose the Bayes optimal model?
</div>

--

- We can specify a loss function. The most common is
.block[
.small[
$$
\begin{split}
L(\hat{\gamma},\gamma) = \boldsymbol{1(\hat{\gamma} \neq \gamma)},
\end{split}
$$
]
]

  that is,
  1. Loss equals zero if the correct model is chosen; and
  2. Loss equals one if incorrect model is chosen.
  
--

- Next, select $\hat{\gamma}$ to minimize Bayes risk. Here, Bayes risk (expected loss over posterior) is
.block[
.small[
$$
\begin{split}
R(\hat{\gamma}) = \sum_{\gamma \in \Gamma} \boldsymbol{1(\hat{\gamma} \neq \gamma)} \cdot \hat{\Pi}_\gamma = 0 \cdot \hat{\Pi}_{\gamma_{\text{true}}} +   \sum_{\gamma \neq \gamma_{\text{true}}} \hat{\Pi}_\gamma = \sum_{\gamma \neq \hat{\gamma}} \hat{\Pi}_\gamma = 1 - \hat{\Pi}_{\hat{\gamma}}
\end{split}
$$
]
]

--

- To minimize $R(\hat{\gamma})$, choose $\hat{\gamma}$ such that $\hat{\Pi}_{\hat{\gamma}}$ is the largest! That is, select the model with the largest posterior probability.



---
## Inference vs prediction

- What if the goal is prediction? Then maybe we should care more about predictive accuracy, rather than selecting specific variables.

--

- For predictions, we care about the posterior predictive distribution, that is
.block[
.small[
$$
\begin{split}
p(y_{n+1}|Y = (y_1, \ldots, y_n)) & = \int_\Gamma \int_{\Theta_\gamma} p(y_{n+1}|\gamma, \theta_\gamma) \cdot \pi(\gamma, \theta_\gamma | Y)  \ \text{d}\theta_\gamma \text{d}\gamma  \\
& = \int_\Gamma \int_{\Theta_\gamma} p(y_{n+1}|\gamma, \theta_\gamma) \cdot \pi(\theta_\gamma | Y, \gamma) \cdot \Pr[\gamma | Y]  \ \text{d}\theta_\gamma \text{d}\gamma  \\
& = \sum_{\gamma \in \Gamma} \int_{\Theta_\gamma} p(y_{n+1}|\gamma, \theta_\gamma) \cdot \pi(\theta_\gamma | Y, \gamma) \cdot \hat{\Pi}_\gamma  \ \text{d}\theta_\gamma \\
& = \sum_{\gamma \in \Gamma} \hat{\Pi}_\gamma \cdot \int_{\Theta_\gamma} p(y_{n+1}|\gamma, \theta_\gamma) \cdot \pi(\theta_\gamma | Y, \gamma)  \ \text{d}\theta_\gamma \\
& = \sum_{\gamma \in \Gamma} \hat{\Pi}_\gamma \cdot p(y_{n+1} | Y, \gamma), \\
\end{split}
$$
]
]

  which is just averaging out the predictions from each model, over all possible models in $\Gamma$, with the posterior probability of each model, and this is known as .hlight[Bayesian model averaging (BMA)].



---
## Bayesian Linear Regression 


** Practical Issues:** the posterior probability that the model is true
.block[
.small[
$$
\begin{split}
\hat{\Pi}_\gamma & = \frac{ \Pi_\gamma p_\gamma(Y) }{ \sum_{\gamma^\star \in \Gamma} \Pi_{\gamma^\star} p_{\gamma^\star}(Y)  }.\\
\end{split}
$$
]
]

--


- We need to calculate marginal likelihoods for ALL models in $\Gamma$ 
  
--


- In general for, we cannot calculate the marginal likelihoods unless we have a proper or conjugate priors  (Normal-Gamma priors within each model)  
  
--

- We need to specify proprer prior distributions on  all common parameters $\theta_\gamma$ in each models!  Conventional priors such as Zellner's g-prior or Ridge Regression to reduce elicitation of prior covariances

--

- Can put priors on hyperparameters cases and integrate or use numerical approximations! 

--

- May not be able to enumerate! Gibbs or MCMC for more flexibility!



---
## Bayesian Variable Selection  (BVS)

- Rewrite each model $\gamma \in \Gamma$ as
.block[
$$\boldsymbol{Y} \mid \alpha, \boldsymbol{\beta}_{\gamma}, \gamma, \phi \sim \mathcal{N}_n(\mathbf{1}_n \alpha + \boldsymbol{X}_{\gamma}\boldsymbol{\beta}_{\gamma}, \phi^{-1}\boldsymbol{I}_{n\times n})$$
]

--

- $\gamma$ represents the set of predictors we want to include in  our model.

--

-  $\gamma = (\gamma_1, \ldots, \gamma_{p}) \in \{0,1\}^p$, so that the cardinality of $\Gamma$ is $2^p$, the number of models in $\Gamma$.

--

$$\gamma_j =  \left\{ \begin{array}{cl} 1 & \text{ if the j'th predictor is included in the model} \\ 0 & \text{ if it is not} \end{array}  \right.$$
 
- $p_\gamma \equiv \sum^p_{j=1} \gamma_j$, so that $p_\gamma$ is the number of predictors included in model $\gamma$
  
- $\boldsymbol{X}_{\gamma}$ ( $n \times p_\gamma$ ) is the matrix of predictors with $\gamma_j = 1$ (wolg design matrix with centered columns)

- $\boldsymbol{\beta}_{\gamma}$ ( $p_\gamma \times 1$ )is the corresponding vector of predictors with $\gamma_j = 1$ 
  


---
## BVS

- Recall that we can also write each model as
.block[
.small[
$$Y_i =  1 \alpha + \boldsymbol{\beta}^T_{\gamma} \boldsymbol{x}_{i\gamma} + \epsilon_i; \ \ \ \  \epsilon_i \overset{iid}{\sim} \mathcal{N}(0, \phi^{-1}).$$
]
]
  
--

- As an example, suppose we had data with 5 potential predictors including the intercept, so that each $\boldsymbol{x}_i = (x_{i1}, x_{i2}, x_{i3},x_{i4},x_{i5})$, and $\boldsymbol{\beta} = (\beta_1, \beta_2, \beta_3, \beta_4, \beta_5)$.

--

- Then for model with $\gamma = (1,0,0,0,0)$, $Y_i = \boldsymbol{\beta}^T_{\gamma} \boldsymbol{x}_{i\gamma} + \epsilon_i$
.block[
.small[
$$\implies Y_i = \alpha + \beta_1 x_{i1} + \epsilon_i; \ \ \ \  \epsilon_i \overset{iid}{\sim} \mathcal{N}(0, 1/\phi),$$
]
]

  with $p_\gamma = 1$.

- Whereas for model with $\gamma = (0,0,1,1,0)$, $Y_i = \alpha +  \boldsymbol{\beta}^T_{\gamma} \boldsymbol{x}_{i\gamma} + \epsilon_i$
.block[
.small[
$$\implies Y_i = \alpha + \beta_3 x_{i3} + \beta_4 x_{i4} + \epsilon_i; \ \ \ \  \epsilon_i \overset{iid}{\sim} \mathcal{N}(0, \sigma^21/\phi),$$
]
]

  with $p_\gamma = 2$.



---
## Steps

The outline for variable selection would be as follows:

1) Write down likelihood under model $\gamma$. That is,
.block[
.small[
$$
\begin{split}
p(\boldsymbol{y} | \boldsymbol{X}, \gamma, \alpha, \boldsymbol{\beta}_{\gamma}, \phi) & \propto (\sigma^2)^{-\frac{n}{2}} \ \textrm{exp} \left\{-\frac{\phi}{2} (\boldsymbol{y} - \mathbf{1} \alpha - \boldsymbol{X}_{\gamma}\boldsymbol{\beta}_{\gamma})^T (\boldsymbol{y} - \mathbf{1} \alpha -\boldsymbol{X}_{\gamma}\boldsymbol{\beta}_{\gamma})\right\}\\
\end{split}
$$
]
]

--

2) Define a prior for $\gamma$, $\Pi_\gamma = \Pr[\gamma]$. 

  + $p(\gamma_j = 1) = .5 \Rightarrow p(\gamma) = .5^p$  Uniform on space of models and $p_\gamma \sim \textsf{Bin}(p, .5)$
  
  + $\gamma_j \mid \pi \overset{iid}{\sim} \textsf{Ber}(\pi)$ and $\pi \sim \textsf{Beta}(a,b)$ then  $p_\gamma \sim \textsf{Beta-Binomial}(a, b)$
  
$$p(p_\gamma \mid p, a, b) = \frac{\Gamma(p + 1) \Gamma(p_\gamma + a) \Gamma(p - p_\gamma + b) \Gamma (a + b) }{\Gamma(p_\gamma+1) \Gamma(p - p_\gamma + 1) \Gamma(p + a + b) \Gamma(a) \Gamma(b)}$$
$$p_\gamma \sim \textsf{Beta-Binomial}(1, 1) \sim \textsf{Unif}(0, p)$$
---
## Prior on model specific parameters

3) Using independent Jeffrey's priors on common parameters and the g-prior  we have

$$\begin{align*}
\pi(\alpha, \phi) & = \phi^{-1}\\
\pi(\boldsymbol{\beta}_{\gamma} | \phi) & = \textsf{N}_p\left(\boldsymbol{\beta}_{0\gamma}= \boldsymbol{0}, \Sigma_{0\gamma} = \frac{g}{\phi} \left[{\boldsymbol{X}_{\gamma}}^T \boldsymbol{X}_{\gamma} \right]^{-1} \right) \\ 
\end{align*}$$




---
## Posteriors

- With those pieces, the conditional posteriors are straightforward 

$$\begin{align*}\alpha \mid \phi, y & \sim \textsf{N}\left(\bar{y}, \frac{1}{n \phi}\right)\\
\boldsymbol{\beta}_{\gamma} \mid \gamma, \phi, g, y &\sim \textsf{N}\left( \frac{g}{1 + g} \hat{\boldsymbol{\beta}}_{\gamma}, \frac{g}{1 + g} \frac{1}{\phi} \left[{\boldsymbol{X}_{\gamma}}^T \boldsymbol{X}_{\gamma} \right]^{-1}  \right) \\
\phi \mid \gamma, y & \sim \textsf{Gamma}( \cdot , \cdot ) \\
p(\gamma \mid y) & \propto p(y \mid \gamma) p(\gamma) 
\end{align*}$$

--
 
- due to conjugacy, the marginal likelihood of $\gamma$ is proportional
to 

$$p(Y \mid \gamma) = C (1 + g)^{\frac{n-p_\gamma -1}{2}} ( 1 + g (1 - R^2_\gamma))^{- \frac{(n-1)}{2}}$$
 
- $R^2_\gamma$ is the usual coefficient of determination for model $\gamma$, 

$$R^2_\gamma = 1 - \frac{(y - \hat{y}_{\gamma})^T(y - \hat{y}_{\gamma})}{(y -  \mathbf{1}\bar{y})^T (y -  \mathbf{1} \bar{y})}$$


--

-  we can run a collapsed Gibbs or MH sampler over just $\Gamma$! 

---
## Summaries

- We can then compute marginal posterior probabilities $\Pr[\gamma | Y]$ for each model and select model with the highest posterior probability.

--

- We can also compute posterior $\Pr[\gamma_j = 1 \mid Y]$, the posterior probability of including the $j$'th predictor, often called .hlight[marginal inclusion probability (MIP)], allowing for uncertainty in the other predictors.

--

- Also straightforward to do model averaging once we all have posterior samples.

--

- The Hoff book works through one example and you can find the Gibbs sampler for doing inference there. I strongly recommend you go through it carefully!

--

- Also paper by Liang et al (2008)  JASA

--
- we will focus on using R packages for implementing


---
##  Examples with BAS
```{r bas}
library(BAS)
data(usair, package="HH")
poll.bma = bas.lm(log(SO2) ~ temp + log(mfgfirms) +
                             log(popn) + wind +
                             precip + raindays,
                  data=usair,
                  prior="g-prior",
                  alpha=nrow(usair), # g = n
                  n.models=2^6,
                  modelprior = uniform(),
                  method="deterministic")
```

---
## Summaries

```{r}
poll.bma
```

---
## Plots of Coefficients

```{r coef_plot, out.width='75%',out.height='75%', fig.height=5,fig.width=8, echo=TRUE}
 beta = coef(poll.bma)
 par(mfrow=c(2,3));  plot(beta, subset=2:7,ask=F)
```

---
##  Summary of Coefficients

```{r}
beta
```

Iterated Expectations!

---
##  Model Space Visualization 

```{r image_plot, out.width='75%',out.height='75%', fig.height=5,fig.width=8, echo=TRUE}

image(poll.bma, rotate=FALSE)
```