---
title: "Lecture 8: More MCMC:  Metropolis-Hastings, Gibbs and Blocking"
author: "Merlise Clyde"
date: "September 26"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(lattice)
require(MCMCpack)
require(hdrcde)
require(coda)
require(mvtnorm)
#require(rethinking)
set.seed(42)
```

## Normal Linear Regression Example

Model
$$\begin{align*}
Y_i \mid \beta, \phi & \overset{iid}{\sim} \textsf{N}(x_i^T\beta, 1/\phi) \\
Y \mid \beta, \phi & \sim \textsf{N}(X \beta, \phi^{-1} I_n) \\
\beta & \sim \textsf{N}(b_0, \Phi_0^{-1}) \\
\phi & \sim \textsf{N}(v_0/2, s_0/2)
\end{align*}$$

--

$x_i$ is a $p \times 1$ vector of predictors and $X$ is $n \times p$ matrix

--

$\beta$ is a $p \times 1$ vector of coefficients

--

$\Phi_0$ is a $p \times p$ prior precision matrix 

--

Multivariate Normal density for $\beta$
$$\pi(\beta \mid b_0, \Phi_0) = \frac{|\Phi_0|^{1/2}}{(2 \pi)^{p/2}}\exp\left\{- \frac{1}{2}(\beta - b_0)^T \Phi_0 (\beta - b_0)  \right\}$$
---
## Full Conditional for $\beta$

$$\begin{align*}
\beta & \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(b_n, \Phi_n^{-1}) \\
b_n & =  (\Phi_0 + \phi X^TX)^{-1}(\Phi_0 b_0  +  \phi X^TX \hat{\beta})\\
\Phi_n & = \Phi_0 + \phi X^TX
\end{align*}$$



---
## Full Conditional for $\phi$

$$ \phi \mid \beta, y_1, \ldots, y_n \sim \textsf{Gamma}((v_0 + n)/2, (s_0 + \sum_i(y_i - x^T_i \beta)))$$



---
##  Choice of Prior Precision

Non-Informative $\Phi_0 \to 0$ 

--

- Formal Posterior given $\phi$
  $$\beta \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(\hat{\beta}, \phi^{-1} (X^TX)^{-1})$$
  
--

- needs $X^TX$ to be full rank for distribution to be unique

## Normal Linear Regression Example

Model
$$\begin{align*}
Y \mid \beta, \phi & \sim \textsf{N}(X \beta, \phi^{-1} I_n) \\
\beta & \sim \textsf{N}(b_0, \Phi_0^{-1}) \\
\phi & \sim \textsf{Gamma}(v_0/2, s_0/2)
\end{align*}$$



- Full Conditional for $\beta$

$$\begin{align*}
\beta & \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(b_n, \Phi_n^{-1}) \\
b_n & =  (\Phi_0 + \phi X^TX)^{-1}(\Phi_0 b_0  +  \phi X^TX \hat{\beta})\\
\Phi_n & = \Phi_0 + \phi X^TX
\end{align*}$$

- Full Conditional for $\phi$

$$ \phi \mid \beta, y_1, \ldots, y_n \sim \textsf{Gamma}((v_0 + n)/2, (s_0 + \sum_i(y_i - x^T_i \beta)))$$

- Choice of Prior Precision $\Phi_0$

---
## Invariance and Choice of Mean/Precision


- What if we transform the $X$ matrix by $\tilde{X} = X H$ where $H$ is $p \times p$ and invertible

--

- obtain the posterior for $\tilde{\beta}$ using $Y$ and $\tilde{X}$  

$$ Y \sim \textsf{N}_n (\tilde{X}\tilde{\beta}, \phi^{-1} I_n)$$

- invariance suggests that since $\tilde{X} \tilde{\beta} = X H  \tilde{\beta} = X \beta$  the posterior for $\beta$ and $H \tilde{\beta}$ should be the same 
(or the posterior of $H^{-1} \beta$
and $\tilde{\beta}$ should be the same)
--

- with some linear algebra we can show that this is true if $b_0 = 0$ and $\Phi_0$ is $k X^TX$ for some $k$  (show!)

---
## Zellner's g-prior

Popular choice is to take $k = \phi/g$ which is a special case of Zellner's g-prior

$$\beta \mid \phi, g \sim \textsf{N}\left(0, \frac{g}{\phi} (X^TX)^{-1}\right)$$

--

- Full conditional 

$$\beta \mid \phi, g \sim \textsf{N}\left(\frac{g}{1 + g} \hat{\beta}, \frac{1}{\phi} \frac{g}{1 + g} (X^TX)^{-1}\right)$$
--

- one parameter $g$ controls shrinkage

--

if $\phi \sim \textsf{Gamma}(v_0/2, s_0/2)$ then posterior is
$$\phi \mid y_1, \ldots, y_n \sim \textsf{Gamma}(v_n/2, s_n/2)$$
--

Conjugate so we could skip Gibbs sampling and sample directly from gamma and then conditional normal!

---
## Ridge Regression 

If $X^TX$ is nearly singular, certain  elements of $\beta$ or (linear combinations of $\beta$) may have huge variances under the $g$-prior (or flat prior) as the MLEs are highly unstable!

--

**Ridge regression** protects against the explosion of variances and ill-conditioning with the conjugate priors:


$$\beta \mid \phi \sim \textsf{N}(0, \frac{1}{\phi \lambda} I_p)$$
--

Posterior for $\beta$  (conjugate case)

$$\beta \mid \phi, \lambda, y_1, \ldots, y_n \sim 
\textsf{N}\left((\lambda I_p + X^TX)^{-1} X^T Y,  \frac{1}{\phi}(\lambda I_p + X^TX)^{-1}
\right)$$




---
##  Bayes Regression

- Posterior mean (or mode) given $\lambda$ is biased, but can show that there **always** is a value of $\lambda$  where the frequentist's expected squared error loss is smaller for the Ridge estimator than MLE!

--

- related to penalized maximum likelihood estimation 

--

-  Choice of $\lambda$

--

-  usual center and standardized $X$ !

--

-  Bayes Regression and choice of $\Phi_0$ in general is a very important problem and provides the foundation  for many variations on shrinkage estimators, variable selection, hierarchical models, nonparameteric regression and more!

--

- Be sure that you can derive the full conditional posteriors for $\beta$ and $\phi$ as well as the joint posterior in the conjugate case!


---
---
## Invariance and Choice of Mean/Precision

-  the model in vector form

$$ Y \sim \textsf{N}_n (X\beta, \phi^{-1} I_n)$$

--

- What if we transform the $X$ matrix by $\tilde{X} = X H$ where $H$ is $p \times p$ and invertible

--

- obtain the posterior for $\tilde{\beta}$ using $Y$ and $\tilde{X}$  

$$ Y \sim \textsf{N}_n (\tilde{X}\tilde{\beta}, \phi^{-1} I_n)$$

- since $\tilde{X} \tilde{\beta} = X H  \tilde{\beta} = X \beta$  invariance suggests that the posterior for $\beta$ and $H \tilde{\beta}$ should be the same 

--
- or the posterior of $H^{-1} \beta$
and $\tilde{\beta}$ should be the same

--

- with some linear algebra we can show that this is true if $b_0 = 0$ and $\Phi_0$ is $k X^TX$ for some $k$  (show!)

---
## Zellner's g-prior

Popular choice is to take $k = \phi/g$ which is a special case of Zellner's g-prior

$$\beta \mid \phi, g \sim \textsf{N}\left(0, \frac{g}{\phi} (X^TX)^{-1}\right)$$

--

- Full conditional 

$$\beta \mid \phi, g \sim \textsf{N}\left(\frac{g}{1 + g} \hat{\beta}, \frac{1}{\phi} \frac{g}{1 + g} (X^TX)^{-1}\right)$$
--

- one parameter $g$ controls shrinkage

--

if $\phi \sim \textsf{Gamma}(v_0/2, s_0/2)$ then posterior is
$$\phi \mid y_1, \ldots, y_n \sim \textsf{Gamma}(v_n/2, s_n/2)$$
--

Conjugate so we could skip Gibbs sampling and sample directly from gamma and then conditional normal!

---
## Ridge Regression 

If $X^TX$ is nearly singular, certain  elements of $\beta$ or (linear combinations of $\beta$) may have huge variances under the $g$-prior (or flat prior) as the MLEs are highly unstable!

--

**Ridge regression** protects against the explosion of variances and ill-conditioning with the conjugate priors:


$$\beta \mid \phi \sim \textsf{N}(0, \frac{1}{\phi \lambda} I_p)$$
--

Posterior for $\beta$  (conjugate case)

$$\beta \mid \phi, \lambda, y_1, \ldots, y_n \sim 
\textsf{N}\left((\lambda I_p + X^TX)^{-1} X^T Y,  \frac{1}{\phi}(\lambda I_p + X^TX)^{-1}
\right)$$




---
##  Bayes Regression

- Posterior mean (or mode) given $\lambda$ is biased, but can show that there **always** is a value of $\lambda$  where the frequentist's expected squared error loss is smaller for the Ridge estimator than MLE!

--

- related to penalized maximum likelihood estimation 

--

-  Choice of $\lambda$

--

-  Bayes Regression and choice of $\Phi_0$ in general is a very important problem and provides the foundation  for many variations on shrinkage estimators, variable selection, hierarchical models, nonparameteric regression and more!

--

- Be sure that you can derive the full conditional posteriors for $\beta$ and $\phi$ as well as the joint posterior in the conjugate case!


---
## Comments



- Why don't we treat each individual $\beta_j$ as a separate block?

--

-  Gibbs always accepts, but can mix slowly if parameters in different blocks are highly correlated!

--

- Use block sizes in Gibbs that are as big as possible to improve  mixing (proven faster convergence)

--

- Collapse the sampler by integrating out as many parameters as possible  (as long as resulting sampler has good mixing)

--

- can use Gibbs steps and (adaptive) Metropolis Hastings steps together

--


-  Introduce latent variables (data augmentation) to allow Gibbs steps  (Next class)





