---
title: "STA 601: Linear Mixed Effects Models"
subtitle: "STA 601 Fall 2021"
author: "Merlise Clyde"
date: "Nov 4, 2021"
header-includes:
  - \include{macros.tex}
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Linear Mixed Effects Models

$$\begin{align*}y_{ij} & = \beta^T x_{ij}+ \gamma ^T z_{ij} + \epsilon_{ij}, \qquad
\epsilon_{ij}  \overset{iid}{\sim}  N(0, \sigma^2) \\
\gamma_j & \overset{iid}{\sim} N_p(0, \Sigma)
\end{align*}$$

--

- Fixed effects contribution $\beta^T x_{ij}$, $x_{ij}$ is a $d \times 1$ vector with $\beta$ is constant across groups $j$, $j = 1, \ldots, J$

--

- Random effects $\gamma^T_j z_{ij}$,  $z_{ij}$ is a $p \times 1$ vector with $\gamma_j \overset{iid}{\sim} N_p(0, \Sigma)$ for groups  $j = 1, \ldots, J$

--

- Designed to accomodate correlated data due to nested/hierarchical structure/repeated measurements

--

- students w/in schools; patients w/in hospitals; additional covariates

--

- As before not inherently Bayesian!   It's just a model/likelihood specification!
  Population parameters, $\theta = (\beta, \Sigma, \sigma^2)$

---
## Likelihoods 

- Complete Data Likelihood $(\{\gamma_i\}, \theta)$

$$L(\{\beta_i\}, \theta)) \propto \prod_j N(\gamma_j; 0, \Sigma) \prod_i N(y_{ij}; \beta^T x_{ij} + \gamma_j^Tz_{ij}, \sigma^2 )$$
--

- Marginal likelihood based on just observed data $(\{y_{ij}\}, \{x_{ij}\}, \{z_{ij}\})$ 

--

$$L(\{\beta_i\}, \theta)) \propto \prod_j \int N(\gamma_j; 0, \Sigma) \prod_i N(y_{ij}; \beta^T x_{ij} + \gamma_j^Tz_{ij}, \sigma^2 ) \, d \gamma_j$$
--

- Option A: we can calculate this integral by brute force algebraically

--

- Option B: (lazy option) We can calculate marginal exploiting properties of Gaussians as sums will be normal -  just read off the first two moments!


---
# Marginal Distribution

- Express observed data as vectors for each group $j$:  $(Y_j, X_j, Z_j)$ where  $Y_j$ is $n_j \times 1$, $X_j$ is $n_j \times d$ and $Z_j$ is $n_j \times p$;

--

- Group Specific Model (1):

$$\begin{align}Y_j  & = X_j \beta + Z_j \gamma + \epsilon_j, \qquad
\epsilon_j  \sim N(0, \sigma^2 I_{n_j})\\
\gamma_j & \overset{iid}{\sim} N(0, \Sigma)
\end{align}$$



--

- Population Mean $E[Y_j] = E[X_j \beta + Z_j \gamma_j + \epsilon_j] = X_j \beta$


--

- Covariance $V[Y_j] = V[X_j \beta + Z_j \gamma_j + \epsilon_j] = Z_j \Sigma Z_j^T + \sigma^2 I_{n_j}$


--

- Group Specific Model (2)
$$Y_j \mid  \beta, \Sigma, \sigma^2 \overset{ind}{\sim}N(X_j \beta, Z_j \Sigma Z_j^T + \sigma^2 I_{n_j})$$


---
## Priors

- Model (1) leads to a simple Gibbs sampler if we use conditional (semi-) conjugate priors on $\theta = (\beta, \Sigma, \phi = 1/\sigma^2)$

$$\begin{align*}
\beta & \sim N(\mu_0, \Psi_0^{-1}) \\
\phi & \sim \textsf{Gamma}(v_0/2, v_o \sigma^2_0/2)
\end{align*}$$

--



- One complication is that the covariance matrix $\Sigma$ must be **positive definite and symmetric**.

--



- "Positive definite" means that for all $x \in \mathcal{R}^p$, $x^T \Sigma x > 0$.

--

- Basically ensures that the diagonal elements of $\Sigma$ (corresponding to the marginal variances) are positive.

--

- Also, ensures that the correlation coefficients for each pair of variables are between -1 and 1.

--

- Our prior for $\Sigma$ should thus assign probability one to set of positive definite matrices.



---
## Inverse-Wishart distribution



- Analogous to the univariate case, the **inverse-Wishart distribution** is the corresponding conditionally conjugate prior for $\Sigma$ (multivariate generalization of the inverse-gamma).

--

- Hoff covers the construction of Wishart and inverse-Wishart random variables in Chapter 7.

--

- A random variable $\Sigma \sim \textrm{IW}_p(\eta_0, \boldsymbol{S}_0^{-1})$, where $\Sigma$ is positive definite and $p \times p$, has pdf

$$p(\Sigma) \propto  \left|\Sigma\right|^{\frac{-(\eta_0 + p + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \textsf{tr}(\boldsymbol{S}_0\Sigma^{-1}) \right\}$$
--

where
 + $\eta_0 > p - 1$ is the "degrees of freedom", and 
 + $\boldsymbol{S}_0$ is a $p \times p$ positive definite matrix.
  
---
## Mean

- For this distribution, $E[\Sigma] = \frac{1}{\eta_0 - p - 1} \boldsymbol{S}_0$, for $\eta_0 > p + 1$.


--

- If we are very confident in a prior guess $\Sigma_0$, for $\Sigma$, then we might set
  + $\eta_0$, the degrees of freedom to be very large, and
  + $\boldsymbol{S}_0 = (\eta_0 - p - 1)\Sigma_0$. 
  
  In this case, $E[\Sigma] = \frac{1}{\eta_0 - p - 1} \boldsymbol{S}_0 = \frac{1}{\eta_0 - p - 1}(\eta_0 - p - 1)\Sigma_0 = \Sigma_0$, and $\Sigma$ is tightly (depending on the value of $\eta_0$) centered around $\Sigma_0$.
  
--

- If we are not at all confident but we still have a prior guess $\Sigma_0$, we might set
  + $\eta_0 = p + 2$, so that the $E[\Sigma] = \frac{1}{\eta_0 - p - 1} \boldsymbol{S}_0$ is finite.
  + $\boldsymbol{S}_0 = \Sigma_0$
  
  
  

---
## Wishart distribution

- Just as we had with the gamma and inverse-gamma relationship in the univariate case, we can also work in terms of the **Wishart distribution** (multivariate generalization of the gamma) instead.

--

- The **Wishart distribution** provides a conditionally-conjugate prior for the precision matrix $\Sigma^{-1}$ in a multivariate normal model.

--

- Specifically, if $\Sigma \sim \textrm{IW}_p(\eta_0, \boldsymbol{S}_0)$, then $\Phi = \Sigma^{-1} \sim \textrm{W}_p(\eta_0, \boldsymbol{S}_0^{-1})$.

--

- A random variable $\Phi \sim \textrm{W}_p(\eta_0, \boldsymbol{S}_0^{-1})$, where $\Phi$ has dimension $(p \times p)$, has pdf

$$\begin{align*}
f(\Phi) \ \propto \ \left|\Phi\right|^{\frac{\eta_0 - p - 1}{2}} \textrm{exp} \left\{-\frac{1}{2} \text{tr}(\boldsymbol{S}_0\Phi) \right\}.
\end{align*}$$


--

- Here, $E[\Phi] = \eta_0 \boldsymbol{S}_0$.


  
---
## Conditional posterior for $\Sigma$

$$\begin{align}Y_j  \mid \beta, \gamma_j, \sigma^2 & \overset{ind}{\sim} N(X_j \beta + Z_j \gamma_j , \sigma^2 I_{n_j})\\
\gamma_j  \mid \Sigma & \overset{iid}{\sim} N(0, \Sigma) \\
\Sigma  & \sim  \textrm{IW}_p(\eta_0, \boldsymbol{S}_0^{-1}) \\
\beta & \sim N(\mu_0, \Psi_0^{-1}) \\
\phi & \sim \textsf{Gamma}(v_0/2, v_o \sigma^2_0/2)
\end{align}$$

- The conditional posterior (full conditional) $\Sigma \mid \boldsymbol{\gamma}, \boldsymbol{Y}$, is then

$$\begin{align*}
\pi(\Sigma & \mid \boldsymbol{\gamma}, \boldsymbol{Y})\propto \pi(\Sigma) \cdot \pi( \boldsymbol{\gamma} \mid \Sigma)\\
& \propto \underbrace{\left|\Sigma\right|^{\frac{-(\eta_0 + p + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \text{tr}(\boldsymbol{S}_0\Sigma^{-1}) \right\}}_{\pi(\Sigma)} \cdot \underbrace{\prod_{j = 1}^{J}\left|\Sigma\right|^{-\frac{1}{2}} \ \textrm{exp} \left\{-\frac{1}{2}\left[\boldsymbol{\gamma}_j^T \Sigma^{-1} \gamma_j\right] \right\}}_{\pi(\boldsymbol{\gamma} \mid \Sigma)}  
\end{align*}$$

--

- $\Sigma \mid \{\gamma_j\}, \boldsymbol{Y} \sim \textrm{IW}_p\left(\eta_0 + J, (\boldsymbol{S}_0+ \sum_{j=1}^J \gamma_j \gamma_j^T)^{-1} \right)$ 
---
## Posterior Continued

$$\begin{align*}
\pi(\Sigma & \mid \boldsymbol{\gamma}, \boldsymbol{Y})\propto \pi(\Sigma) \cdot \pi( \boldsymbol{\gamma} \mid \Sigma)\\
& \propto \left|\Sigma\right|^{\frac{-(\eta_0 + p + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \text{tr}(\boldsymbol{S}_0\Sigma^{-1}) \right\} \cdot \prod_{j = 1}^{J}\left|\Sigma\right|^{-\frac{1}{2}} \ \textrm{exp} \left\{-\frac{1}{2}\left[\boldsymbol{\gamma}_j^T \Sigma^{-1} \gamma_j\right] \right\}  \\
& \propto \left|\Sigma\right|^{\frac{-(\eta_0 + p + J + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \left[\text{tr}\left[\boldsymbol{S}_0\Sigma^{-1} \right] +\sum_{j=1}^J\gamma_j^T \Sigma^{-1} \gamma_j \right]\right\} ,\\
\\
& \propto \left|\Sigma\right|^{\frac{-(\eta_0 + p + J + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \left[\text{tr}\left[\boldsymbol{S}_0\Sigma^{-1} \right] +\sum_{j=1}^J \text{tr}\left[\gamma_j \gamma_j^T \Sigma^{-1} \right]\right]\right\} ,\\
\\
& \propto \left|\Sigma\right|^{\frac{-(\eta_0 + p + J + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \text{tr}\left[\boldsymbol{S}_0\Sigma^{-1}  +\sum_{j=1}^J \gamma_j \gamma_j^T \Sigma^{-1} \right]\right\} ,\\
\\
& \propto \left|\Sigma\right|^{\frac{-(\eta_0 + p + J + 1)}{2}} \textrm{exp} \left\{-\frac{1}{2} \text{tr}\left[ \left(\boldsymbol{S}_0  +\sum_{j=1}^J \gamma_j \gamma_j^T \right)\Sigma^{-1} \right]\right\} ,\\
\\
\end{align*}$$


---
## MCMC Sampling

- Model (1) leads to a simple Gibbs sampler if we use conditional (semi-) conjugate priors on $\theta = (\beta, \Sigma, \phi = 1/\sigma^2)$

--

- Model (2) can be challenging to update the variaance components!  no conjugacy and need to ensure that MH updates maintain the positive-definiteness of $\Sigma$  (can reparameterize)


--

- Is Gibbs always more efficient?

--

- No -  because the Gibbs sampler can have high autocorrelation in updating the $\{\gamma_j \}$ from their full conditional and then updating $\theta$ from their full full conditionals given the $\{ \gamma_j\}$

--

- slow mixing

--
- update $\beta$ using (2) instead of (1)  (marginalization so is independent of $\gamma_j$'s


---
##  Marginal update for $\beta$

$$\begin{align*}Y_j \mid  \beta, \Sigma, \sigma^2 & \overset{ind}{\sim}N(X_j \beta, Z_j \Sigma Z_j^T + \sigma^2 I_{n_j}) \\
\beta & \sim N(\mu_0, \Psi_0^{-1})
\end{align*}$$

--
- Let $\Phi_j = (Z_j \Sigma Z_j^T + \sigma^2 I_{n_j})^{-1}$  (precision in model 2)

--

$$\begin{align*}
\pi(\beta  & \mid \Sigma, \sigma^2, \boldsymbol{Y})  \propto |\Psi_0|^{1/2} 
\exp\left\{- \frac{1}{2} (\beta - \mu_0)^T \Psi_0 (\beta - \mu_0)\right\} \cdot \\
& \qquad \qquad \qquad\prod_{j=1}^{J} |\Phi_j|^{1/2} \exp \left\{ - \frac{1}{2} (Y_j - X_j \beta)^T \Phi_j (Y_j - X_j \beta ) \right\} \\
\\
& \propto \exp\left\{- \frac{1}{2} \left( (\beta - \mu_0)^T \Psi_0 (\beta - \mu_0) +
\sum_j (Y_j - X_j \beta)^T \Phi_j (Y_j - X_j \beta ) \right) \right\}
\end{align*}$$

---
## Marginal Posterior for $\beta$

$$\begin{align*}
\pi(\beta  & \mid \Sigma, \sigma^2, \boldsymbol{Y})  \\
& \propto \exp\left\{- \frac{1}{2} \left( (\beta - \mu_0)^T \Psi_0 (\beta - \mu_0) +
\sum_j (Y_j - X_j \beta)^T \Phi_j (Y_j - X_j \beta ) \right) \right\}
\end{align*}$$

--

- Expand, read off precision and fix up mean so that it is a function of MLE's

--

- precision

$$\Psi_n = \Psi_0 + \sum_{j=1}^J X_j^T \Phi_j X_j$$


- mean 

$$\mu_n = \left(\Psi_0 + \sum_{j=1}^J X_j^T \Phi_j X_j\right)^{-1} \left(\Psi_0 \mu_0 + 
\sum_{j=1}^J X_j^T \Phi_j X_j \hat{\beta}_j\right)$$

where 
$\hat{\beta}_j = (X_j^T \Phi X_j)^{-1} X_j^T \Phi_j Y_j\right)$ is the generalized least squares estimate of $\beta$ for group $j$


---
## Full conditional for $\sigma^2$ or $\phi$

$$\begin{align}Y_j  \mid \beta, \gamma_j, \sigma^2 & \overset{ind}{\sim} N(X_j \beta + Z_j \gamma_j , \sigma^2 I_{n_j})\\
\gamma_j  \mid \Sigma & \overset{iid}{\sim} N(0, \Sigma) \\
\Sigma  & \sim  \textrm{IW}_p(\eta_0, \boldsymbol{S}_0^{-1}) \\
\beta & \sim N(\mu_0, \Psi_0^{-1}) \\
\phi & \sim \textsf{Gamma}(v_0/2, v_o \sigma^2_0/2)
\end{align}$$

$$\pi(\phi \mid \beta, \{\gamma_j\} \{Y_j\}) \propto \textsf{Gamma}(\phi; v_0/2, v_o \sigma^2_0/2) \prod_j N(Y_j; X_j \beta + Z_j \gamma_j , \phi^{-1} I_{n_j}))$$
$$\phi \mid \{Y_j \}, \beta, \{\gamma_j\} \sim \textsf{Gamma}\left(\frac{v_0 + \sum_j n_j}{2}, \frac{v_o \sigma^2_0  + \sum_j \|Y_j - X_j\beta - Z_j\gamma_j \|^2}{2}\right)$$
---
## Full conditional for $\{ \gamma_j \}$

$$\begin{align}Y_j  \mid \beta, \gamma_j, \sigma^2 & \overset{ind}{\sim} N(X_j \beta + Z_j \gamma_j , \sigma^2 I_{n_j})\\
\gamma_j  \mid \Sigma & \overset{iid}{\sim} N(0, \Sigma) \\
\Sigma  & \sim  \textrm{IW}_p(\eta_0, \boldsymbol{S}_0^{-1}) \\
\beta & \sim N(\mu_0, \Psi_0^{-1}) \\
\phi & \sim \textsf{Gamma}(v_0/2, v_o \sigma^2_0/2)
\end{align}$$

$$\pi(\gamma \mid \beta, \phi, \Sigma) \propto \textsf{N}(\gamma_j; 0, \Sigma) \prod_j N(Y_j; X_j \beta + Z_j \gamma_j , \phi^{-1} I_{n_j}))$$

