---
title: "STA 360/602L: Module 8.1"
subtitle: "The multinomial model"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(Compositional)
library(MCMCpack)
library(gtools)
```


## Categorical data (univariate)

- Suppose

  + $Y \in \{1,\ldots, D\}$;
  
  + $\Pr(y = d) = \theta_d$ for each $d = 1,\ldots, D$; and
  
  + $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_D)$.
  
- Then the pmf of $Y$ is
.block[
.small[
$$\Pr[y = d| \boldsymbol{\theta}] = \prod_{d=1}^D \theta_d^{\mathbb{1}[y = d]}.$$
]
]

--

- We say $Y$ has a .hlight[multinomial distribution] with sample size 1, or a .hlight[categorical distribution].

--

- Write as $Y | \boldsymbol{\theta} \sim \textrm{Multinomial}(1,\boldsymbol{\theta})$ or $Y | \boldsymbol{\theta} \sim \textrm{Categorical}(\boldsymbol{\theta})$.

--

- Clearly, this is just an extension of the Bernoulli distribution.




---
## Dirichlet distribution

- Since the elements of the probability vector $\boldsymbol{\theta}$ must always sum to one, the support is often called a .hlight[simplex].

--

- A conjugate prior for categorical/multinomial data is the .hlight[Dirichlet distribution].

--

- A random variable $\boldsymbol{\theta}$ has a .hlight[Dirichlet distribution] with parameter $\boldsymbol{\alpha}$, if
.block[
.small[
$$p[\boldsymbol{\theta} | \boldsymbol{\alpha}] = \dfrac{\Gamma\left(\sum_{d=1}^D \alpha_d\right)}{\prod_{d=1}^D \Gamma(\alpha_d)} \prod_{d=1}^D \theta_d^{\alpha_d-1}, \ \ \ \alpha_d > 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]

  where $\boldsymbol{\alpha} = (\alpha_1,\ldots,\alpha_D)$, and
.block[
.small[
$$\sum_{d=1}^D \theta_d = 1, \ \ \theta_d \geq 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]

--

- We write this as $\boldsymbol{\theta} \sim \textrm{Dirichlet}(\boldsymbol{\alpha}) = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_D)$.

--

- The Dirichlet distribution is a multivariate generalization of the .hlight[beta distribution].



---
## Dirichlet distribution

- Write
.block[
.small[
$$\alpha_0 = \sum_{d=1}^D \alpha_d \ \ \ \textrm{and} \ \ \ \alpha_d^\star = \dfrac{\alpha_d}{\alpha_0}.$$
]
]

--

- Then we can re-write the pdf slightly as
.block[
.small[
$$p[\boldsymbol{\theta} | \boldsymbol{\alpha}] = \dfrac{\Gamma\left(\alpha_0\right)}{\prod_{d=1}^D \Gamma(\alpha_d)} \prod_{d=1}^D \theta_d^{\alpha_d-1}, \ \ \ \alpha_d > 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]
  
--

- Properties:

  - 
.block[
.small[
$$\mathbb{E}[\theta_d] = \alpha_d^\star;$$
]
]

--

  - 
.block[
.small[
$$\textrm{Mode}[\theta_d] = \dfrac{\alpha_d - 1}{\alpha_0 - d};$$
]
]

--
  
  - 
.block[
.small[
$$\mathbb{Var}[\theta_d] = \dfrac{\alpha_d^\star(1-\alpha_d^\star)}{\alpha_0 + 1} = \dfrac{\mathbb{E}[\theta_d](1-\mathbb{E}[\theta_d])}{\alpha_0 + 1};$$
]
]
  
--

  - 
.block[
.small[
$$\mathbb{Cov}[\theta_d,\theta_k] = \dfrac{\alpha_d^\star\alpha_k^\star}{\alpha_0 + 1} = \dfrac{\mathbb{E}[\theta_d]\mathbb{E}[\theta_k]}{\alpha_0 + 1}.$$
]
]





---
## Dirichlet examples

$\textrm{Dirichlet}(1,1,1)$

```{r fig.height=4.7, echo=F}
draws <- rdirichlet(1000, c(1,1,1))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(10,10,10)$

```{r fig.height=4.7, echo=F}
draws <- rdirichlet(1000, c(10,10,10))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(100,100,100)$

```{r fig.height=4.7, echo=F}
draws <- rdirichlet(1000, c(100,100,100))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(1,10,1)$

```{r fig.height=4.7, echo=F}
draws <- rdirichlet(1000, c(1,10,1))
bivt.contour(draws)
```




---
## Dirichlet examples

$\textrm{Dirichlet}(50,100,10)$

```{r fig.height=4.7, echo=F}
draws <- rdirichlet(1000, c(50,100,10))
bivt.contour(draws)
```





---
## Likelihood

- Let $Y_i, \ldots, Y_n | \boldsymbol{\theta} \sim \textrm{Categorical}(\boldsymbol{\theta})$.

--

- Recall
.block[
.small[
$$\Pr[y_i = d| \boldsymbol{\theta}] = \prod_{d=1}^D \theta_d^{\mathbb{1}[y_i = d]}.$$
]
]

--

- Then,
.block[
.small[
$$
\begin{split}
p[Y | \boldsymbol{\theta}] = p[y_1,\ldots,y_n | \boldsymbol{\theta}] = \prod_{i=1}^n \prod_{d=1}^D \theta_d^{\mathbb{1}[y_i = d]} = \prod_{d=1}^D  \theta_d^{\sum_{i=1}^n \mathbb{1}[y_i = d]} = \prod_{d=1}^D  \theta_d^{n_d}
\end{split}
$$
]
]

  where $n_d$ is just the number of individuals in category $d$.

--

- Maximum likelihood estimate of $\theta_d$ is
.block[
.small[
$$\hat{\theta}_d = \dfrac{n_d}{n}, \ \ d = 1,\ldots, D$$
]
]



---
## Posterior

- Set $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_D)$.
.block[
.small[
$$
\begin{split}
\pi(\boldsymbol{\theta} | Y) & \propto p[Y| \boldsymbol{\theta}] \cdot \pi[\boldsymbol{\theta}]\\
& \propto \prod_{d=1}^D  \theta_d^{n_d} \prod_{d=1}^D \theta_d^{\alpha_d-1} \\
& \propto \prod_{d=1}^D  \theta_d^{\alpha_d + n_d - 1}\\
& = \textrm{Dirichlet}(\alpha_1 + n_1,\ldots,\alpha_D + n_D)
\end{split}
$$
]
]

--

- Posterior expectation:
.block[
.small[
$$\mathbb{E}[\theta_d | Y] = \dfrac{\alpha_d + n_d}{\sum_{d^\star=1}^D (\alpha_{d^\star} + n_{d^\star})}.$$
]
]



---
## Combining information

- For the prior, we have
.block[
.small[
$$\mathbb{E}[\theta_d] = \dfrac{\alpha_d}{\sum_{d^\star=1}^D \alpha_{d^\star}}$$
]
]

--

- We can think of 

  + $\theta_{0d} = \mathbb{E}[\theta_d]$ as being our **"prior guess"** about $\theta_d$, and
  
  + $n_0 = \sum_{d^\star=1}^D \alpha_{d^\star}$ as being our **"prior sample size"**.
  
  

--

- We can then rewrite the prior as $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(n_0\theta_{01},\ldots,n_0\theta_{0D})$.




---
## Combining information

- We can write the posterior expectation as:
.block[
.small[
$$
\begin{split}
\mathbb{E}[\theta_d | Y] & = \dfrac{\alpha_d + n_d}{\sum_{d^\star=1}^D (\alpha_{d^\star} + n_{d^\star})}\\
& = \dfrac{\alpha_d}{\sum_{d^\star=1}^D \alpha_{d^\star} + \sum_{d^\star=1}^D n_{d^\star}} + \dfrac{n_d}{\sum_{d^\star=1}^D \alpha_{d^\star} + \sum_{d^\star=1}^D n_{d^\star}}\\
& = \dfrac{n_0\theta_{0d}}{n_0 + n} + \dfrac{n \hat{\theta}_d}{n_0 + n}\\
& = \dfrac{n_0}{n_0 + n} \theta_{0d} + \dfrac{n}{n_0 + n} \hat{\theta}_d.
\end{split}
$$
]
]

  since
MLE is
.block[
.small[
$$\hat{\theta}_d = \dfrac{n_d}{n}$$
]
]

--

- Once again, we can express our posterior expectation as a weighted average of the prior expectation and MLE.

--

- <div class="question">
We can also extend the Dirichlet-multinomial model to more variables (contingency tables).
</div>




---
## Example: pre-election polling

- Fox News Nov 3-6 pre-election survey of 1295 likely voters for the 2016 election.

--

- For those interested, [FiveThirtyEight](https://projects.fivethirtyeight.com/) is an interesting source for pre-election polls.

--

- Out of 1295 respondents, 622 indicated support for Clinton, 570 for Trump, and the remaining 103 for other candidates or no opinion.

--

- Drawing inference from pre-election polls is way more complicated and nuanced that this. We only use the data here for this simple illustration.

--

- Assuming no other information on the respondents, we can assume simple random sampling and use a multinomial distribution with parameter $\boldsymbol{\theta} = (\theta_1,\theta_2,\theta_3)$, the proportion, in the survey population, of Clinton supporters, Trump supporters and other candidates or no opinion.



---
## Example: pre-election polling

- With a noninformative uniform prior, we have $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(1,1,1)$.

--

- The resulting posterior is $\textrm{Dirichlet}(1 + n_1,1 + n_2,1 + n_3) = \textrm{Dirichlet}(623,571,104)$.

--

- Suppose we wish to compare the proportion of people who would vote for Trump versus Clinton, we could examine the posterior distribution of $\theta_1-\theta_2$.

--

- We can even compute the probability $\Pr(\theta_1 > \theta_2 | Y)$.


---
## Example: pre-election polling

```{r fig.height=3.8}
#library(gtools)
PostSamples <- rdirichlet(10000, alpha=c(623,571,104))
#dim(PostSamples)
hist((PostSamples[,1] - PostSamples[,2]),col=rainbow(20),xlab=expression(theta[1]-theta[2]),
     ylab="Posterior Density",freq=F,breaks=50,
     main=expression(paste("Posterior density of ",theta[1]-theta[2])))
```



---
## Example: pre-election polling

- Posterior probability that Clinton had more support than Trump in the survey population, that is, $\Pr(\theta_1 > \theta_2 | Y)$, is

  ```{r fig.height=3.5}
#library(gtools)
mean(PostSamples[,1] > PostSamples[,2])
```

--

- Once again, this is just a simple illustration with a very small subset of the 2016 pre-election polling data.

--

- Inference for pre-election polls is way more complex and nuanced that this.




---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




