---
title: "STA 360/602L: Module 8.2"
subtitle: "Finite mixture models: univariate categorical data"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Multinomial model recap

- Suppose $y_i, \ldots, y_n | \boldsymbol{\theta} \overset{iid}{\sim} \textrm{Categorical}(\boldsymbol{\theta})$, then
.block[
.small[
$$\Pr[y_i = d| \boldsymbol{\theta}] = \prod_{d=1}^D \theta_d^{\mathbb{1}[y_i = d]},$$
]
]

--

- With prior $\pi[\boldsymbol{\theta}] = \textrm{Dirichlet}(\boldsymbol{\alpha})$, we have
.block[
.small[
$$\pi[\boldsymbol{\theta}] \propto \prod_{d=1}^D \theta_j^{\alpha_j-1}, \ \ \ \alpha_j > 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]
  
--

- So that the posterior is
.block[
.small[
$$
\begin{split}
\pi(\boldsymbol{\theta} | Y) = \textrm{Dirichlet}(\alpha_1 + n_1,\ldots,\alpha_D + n_D)
\end{split}
$$
]
]

--

- However, what if our data actually comes from $K$ different sub-populations of groups of people?

--

- For example, if our data comes from men and women, and we don't expect marginal independence across the two groups (vote turnout, income, etc), then we have a mixture of distributions.



---
## Finite mixture of multinomials

- With our data coming from a "combination" or "mixture" of sub-populations, we no longer have independence across all observations, so that the likelihood $p[Y| \boldsymbol{\theta}] \neq \prod\limits_{i=1}^n \prod\limits_{d=1}^D \theta_j^{\mathbb{1}[y_i = d]}$.

--

- However, we can still have "conditional independence" within each group.

--

- Unfortunately, we do not always know the indexes for those groups.

--

- That is, we know our data contains $K$ different groups, but we actually do not know which observations belong to which groups.

--

- **Solution**: introduce a .hlight[latent variable] $z_i$ representing the group/cluster indicator for each observation $i$, so that each $z_i \in \{1,\ldots, K\}$.

--

- This is a form of .hlight[data augmentation], but we will define that properly later.


---
## Finite mixture of multinomials

- Given the cluster indicator $z_i$ for observation $i$, write
  + $\Pr(y_i = d | z_i) = \psi_{z_i,d} \equiv \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]}$, and
  + $\Pr(z_i = k) = \lambda_k \equiv \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]}$.
  
--

- Then, the marginal probabilities we care about will be
.block[
.small[
$$
\begin{split}
\theta_d & = \Pr(y_i = d)\\
& = \sum_{k=1}^K \Pr(y_i = d | z_i = k) \cdot \Pr(z_i = k)\\
& = \sum_{k=1}^K \lambda_k \cdot \psi_{k,d}, \\
\end{split}
$$
]
]

  which is a .hlight[finite mixture of multinomials], with the weights given by $\lambda_k$.
  
<!-- -- -->

<!-- - Intuitively, if we randomly select any $y_i$, the probability that the $y_i$ is equal to $d$, $\Pr(y_i = d)$, is a weighted average of the probability of $y_i$ is equal to $d$ within each cluster/group. -->




---
## Posterior inference

- Write

  + $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_K)$, and
  
--

  + $\boldsymbol{\psi} = \{\psi_{z_i,d}\}$ to be a $K \times D$ matrix of probabilities, where each $k$th row is the vector of probabilities for cluster $k$.

--

- The observed data likelihood is
.block[
.small[
$$
\begin{split}
p\left[Y = (y_1, \ldots, y_n) | Z = (z_1, \ldots, z_n), \boldsymbol{\psi}, \boldsymbol{\lambda}\right] & = \prod_{i=1}^n \prod\limits_{d=1}^D \Pr\left(y_i = d | z_i, \psi_{z_i,d} \right)\\
& = \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]},
\end{split}
$$
]
]

  which includes products (and not the sums in the mixture pdf), and as you will see, makes sampling a bit easier.
  
--

- Next we need priors.



---
## Posterior inference

- First, for $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_K)$, the vector of cluster probabilities, we can use a Dirichlet prior. That is,
.block[
.small[
$$\pi[\boldsymbol{\lambda}] = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_K) \propto \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1}.$$
]
]
  
--

- For $\boldsymbol{\psi}$, we can assume independent Dirichlet priors for each cluster vector $\boldsymbol{\psi}_k = (\psi_{k,1}, \ldots, \psi_{k,D})$. That is, for each $k = 1, \ldots, K$, 
.block[
.small[
$$\pi[\boldsymbol{\psi}_k] = \textrm{Dirichlet}(a_1,\ldots,a_d) \propto \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1}.$$
]
]
  
--

- Finally, from our distribution on the $z_i$'s, we have
.block[
.small[
$$
\begin{split}
p\left[Z = (z_1, \ldots, z_n) | \boldsymbol{\lambda}\right] & = \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]}.
\end{split}
$$
]
]




---
## Posterior inference

- Note that the unobserved variables and parameters are $Z = (z_1, \ldots, z_n)$, $\boldsymbol{\psi}$, and $\boldsymbol{\lambda}$.

--

- So, the joint posterior is
.block[
.small[
$$
\begin{split}
\pi\left(Z, \boldsymbol{\psi}, \boldsymbol{\lambda} | Y \right) & \propto p\left[Y | Z, \boldsymbol{\psi}, \boldsymbol{\lambda}\right]  \cdot p(Z| \boldsymbol{\psi}, \boldsymbol{\lambda})  \cdot \pi(\boldsymbol{\psi}, \boldsymbol{\lambda}) \\
\\
& \propto \left[ \prod_{i=1}^n \prod\limits_{d=1}^D p\left(y_i = d | z_i, \psi_{z_i,d} \right) \right]  \cdot p(Z| \boldsymbol{\lambda})  \cdot \pi(\boldsymbol{\psi}) \cdot \pi(\boldsymbol{\lambda}) \\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]} \right) \\
& \ \ \ \ \ \times \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \\
& \ \ \ \ \ \times \left( \prod\limits_{k=1}^K \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \\
& \ \ \ \ \ \times \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right). \\
\end{split}
$$
]
]



---
## Posterior inference

- First, we need to sample the $z_i$'s, one at a time, from their full conditionals. 

--

- For $i = 1, \ldots, n$, sample $z_i \in \{1,\ldots, K\}$ from a categorical distribution (multinomial distribution with sample size one) with probabilities
.block[
.small[
$$
\begin{split}
\Pr[z_i = k | \dots ] & = \Pr[z_i = k | y_i, \boldsymbol{\psi}_k, \lambda_k] \\
\\
& = \dfrac{ \Pr[y_i, z_i = k | \boldsymbol{\psi}_k, \lambda_k] }{ \sum\limits^K_{l=1} \Pr[y_i, z_i = l | \boldsymbol{\psi}_l, \lambda_l] } \\
\\
& = \dfrac{ \Pr[y_i | z_i = k, \boldsymbol{\psi}_k] \cdot \Pr[z_i = k, \lambda_k] }{ \sum\limits^K_{l=1} \Pr[y_i | z_i = l, \boldsymbol{\psi}_l] \cdot \Pr[z_i = l, \lambda_l] } \\
\\
& = \dfrac{ \psi_{k,d} \cdot \lambda_k }{ \sum\limits^K_{l=1} \psi_{l,d} \cdot \lambda_l }. \\
\end{split}
$$
]
]



---
## Posterior inference

- Next, sample each cluster vector $\boldsymbol{\psi}_k = (\psi_{k,1}, \ldots, \psi_{k,D})$ from
.block[
.small[
$$
\begin{split}
\pi[\boldsymbol{\psi}_k | \dots ] & \propto \pi\left(Z, \boldsymbol{\psi}, \boldsymbol{\lambda} | Y \right) \\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]} \right) \cdot \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \cdot \left( \prod\limits_{k=1}^K \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left(  \prod\limits_{d=1}^D \psi_{k,d}^{n_{k,d}} \right) \cdot \left( \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \\
\\
& = \left(  \prod\limits_{d=1}^D \psi_{k,d}^{a_d + n_{k,d} - 1} \right) \\
\\
& \equiv \textrm{Dirichlet}\left(a_1 + n_{k,1},\ldots,a_D + n_{k,D}\right).
\end{split}
$$
]
]

  where $n_{k,d} = \sum\limits_{i: z_i = k} \mathbb{1}[y_i = d]$, the number of individuals in cluster $k$ that are assigned to category $d$ of the levels of $y$.



---
## Posterior inference

- Finally, sample $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_K)$, the vector of cluster probabilities from
.block[
.small[
$$
\begin{split}
\pi[\boldsymbol{\lambda} | \dots ] & \propto \pi\left(Z, \boldsymbol{\psi}, \boldsymbol{\lambda} | Y \right) \\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]} \right) \cdot \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \cdot \left( \prod\limits_{k=1}^K \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left( \prod\limits_{k=1}^K \lambda_k^{n_k} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k + n_k - 1} \right)\\
\\
& \equiv \textrm{Dirichlet}\left(\alpha_1 + n_1,\ldots,\alpha_K + n_K\right),\\
\end{split}
$$
]
]

  where $n_k = \sum\limits_{i=1}^n \mathbb{1}[z_i = k]$, the number of individuals assigned to cluster $k$.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




